### 2022-03-18 paper.tex
my greek drama final paper

% tell em what you're gonna tell em

In this paper, I will present a reading Sophocles' \textit{Ajax} through the lens of the
Peloponnesian war. I will justify that lens based on the play's historical context at the
time it was conceived, and based on Sophocles' life and experiences. I will demonstrate
that the reading allows us to understand \textit{Ajax}'s treatment of the goddess Athena,
of militarism and imperialism as a way of life, and of a debate between autocratic and
democratic constitutions.

% why would i think to read Ajax through this lens?

%% yes, i know that ajax predates the war.

In justifying this lens, I must, most importantly, vie with the fact that \textit{Ajax}
was almost certainly written before the Peloponnesian war proper. Though dating Sophocles'
plays is a fraught subject, it is generally accepted that \textit{Ajax} is either the
earliest or among the earliest of the surviving dramas \footcites[\enquote{Dating the
Plays}]{meineck_four}[p. 38]{bates_soph}. Recent estimates place the first performance
of Ajax in the 440s B.C.E. \footcite[p. 10]{finglass_ajax}. At a first glance, this would
place it well before the Peloponnesian war, which is reliably dated to have begun in 431
B.C.E. \footcite[p. 32]{lazenby_war}.

My interest is not, however, in the bloody boots-on-the-ground military conflict to which
Sophocles' later plays respond; rather, in the ongoing tensions between Athenian imperial
expansion and Sparta's more conservative holdings which sparked the war. Thucydides
describes a near-constant conflict between the Athenians and the Spartans beginning almost
immediately after the Persian War \footcite[ch. 1.18]{thucydides_war}. Thucydides, himself
an Athenian who wrote before, during and shortly after the war, is eager to acknowledge
that the war was caused by the Spartans' fear of, and desire to contain, Athens' imperial
expansion  \footcite[ch. 1.23]{thucydides_war}. (An investigation of the social and
political factors which led Thucydides to publish that claim would be fascinating, but
beyond the scope of this paper.)

%% TODO: what the hell is going on in athens at this time?

%% FORESHADOW: sophocles foresaw the war, and ajax is his way of warning his peers of the
%% horrors to come.

%% well, thucydides all but tells me to compare the peloponnesian war w/ the trojan

Thucydides' \textit{History of the Peloponnesian War} also invites this lens by devoting
much of his first book to comparing the Peloponnesian war with the Trojan
war \footcite[e.g. ch. 1.23]{thucydides_war}. If we accept that Sophocles was aware of the
rising tensions between the Athenians and the Spartans when he wrote \textit{Ajax}, then
it is not difficult to imagine him drawing the same parallel as Thucydides between a
contemporary conflict and the Greeks' most historicized mythic war. But we need not merely
imagine, since once one looks through the lens of (pre-) Peloponnesian War Hellenic
geopolitics, a number of otherwise murky features in the text become clear.

% demonstrate the lens: Autocracy and democracy

\textit{Ajax} contains some rather blunt attacks against the coercive nature of autocracy,
and at least one more subtle reference to the voluntary civic engagement required for
democracy. The ideological conflict between autocracy and democracy drove the
Peloponnesian war (or at least, was used to justify it), but categorically could not have
existed during the Trojan War, as democracy was not invented for several centuries.

%% Menelaus

The two least sympathetic characters in \textit{Ajax}, Menelaus and Agamemnon, both
espouse (straw men of) autocracy. Menelaus pronounces, \textquote{Now it is, in truth, the
mark of a base nature when a commoner does not think it right to obey those who stand over
him. Never can the laws maintain a prosperous course in a city where fear has no fixed
place} \footcite[ll. 1072--1075]{ajax}. In this moment, Sophocles has a Spartan advocate
for a rigid social hierarchy without upward mobility, held in place by fear. Menelaus is
also doing this in the same breath as ordering the desecration of the corpse of a
traumatized war hero. Spoken before a crowd of Athenian men, most of them
\enquote{commoners,} and all of them soldiers or veterans, this must have been a powerful
denouncement of autocracy.

%% Aga 1 - irreverent

For Agamemnon, not just a Spartan but their king, Sophocles is even more explicit, having
him say, \textquote{Reverence, I tell you, is not easily practiced by the
autocrat} \footcite[l. 1350]{ajax}. For a second time in \textit{Ajax}, Sophocles tells
us that autocratic rule precludes, or at least devalues, just religious observance. In the
deeply religious society of Hellenic Greece, Sophocles associates autocracy with
godlessness and wickedness, and thus implicitly joins democracy with piety.

%% Aga 2 - shut up & listen

In the same moment, Agamemnon also resists the wisdom of Odysseus, telling him,
\textquote{A good man should listen to those in
charge} \footcite[l. 1352]{ajax}. Odysseus is not entirely sympathetic in \textit{Ajax},
but he is still presented as a paragon of wisdom, piety and morality, at least when
juxtaposed with Agamemnon. And yet Sophocles shows Agamemnon arguing with Odysseus for no
purpose other than personal vendetta, and with no more of an argument than \enquote{you
should listen to me because I am king.} This critique of aristocracy, that powerful rulers
become so rigid and self-assured that they ignore good advice and therefore enact bad
policies, is typical of both classical and modern democratic rhetoric.

%% ajax - a democratic citizen

Sophocles presents Ajax's involvement in the Trojan war as the willing participation of an
outsider in democracy. When defending Ajax against Menelaus' accusations of
insubordination, Teucer says, \textquote{Did he not sail of his own accord? As his own
master? On what grounds are you his commander? \textelp{} Nowhere was it established among
your lawful powers that you should order him any more than he
you} \footcite[ll. 1099--1104]{ajax}. This defense of personal sovereignty, that each
citizen/soldier chose to participate in and contribute to society/the war effort freely,
without duress, and without binding themselves to the will of an individual, could have
come from the textbook of my freshman gen ed course on mixed constitutions. And to
Sophocles, this is what makes Ajax noble: he makes his choices for no reason except that
he thinks they are right, and he collaborates with others to achieve larger goals, even
when personal or ideological differences lead to disagreement.
%% CONSIDER: tie this back in to ``all this doesn't make sense in the actual trojan war,
%% bc a. democracy didn't exist yet, and b. thucydides tells us that agamemnon totally did
%% coerce all his allies''

The ideological conflict between autocracy and democracy justifies this lens, in that the
Peloponnesian war explains easily-recognizable features of the text which make little
sense without the political context of Sophocles' time, but it is not terribly exciting to
learn that a Sophocles play promotes democracy and decries autocracy. To learn anything
new, we must focus our lens on Sophocles' treatment of Athenian militarism and
imperialism, and of the city's god, Athena.

% gain insight from the lens:
% athena - what's her deal?

%% is she the noble goddess of wisdom?

\textit{Ajax} is absolutely rife with references to Athena, who seems to play two somewhat
distinct roles. On the one hand, she is the patron of Odysseus and defender of the Danaans
(who, through this lens, must be taken to represent the Athenians in Sophocles'
time). Odysseus tells us that Athena is \textquote{dearest to \textins{him} of the
  gods} \footcite[l. 14]{ajax}, and later in the play, the chorus praises him for his
wisdom \footcite[ll. 1374--1375]{ajax}, a trait typically associated with Athena. His
Athenian wisdom allows Odysseus to see past the hatred which has blinded Menelaus and
Agamemnon, and to resolve the drama of \textit{Ajax} in an honorable, productive
way.

Athena also tells us that it was her intervention which prevented Ajax from slaughtering
the Greek army, reducing the impact of his crimes to a handful of sheep and herdsmen
\footcite[ll. 40--75]{ajax}. If we read Athena as a personification of the city, we get a
picture of Athens as the imperial protector, using its wisdom to avert conflicts and
catastrophes among its subjects. It's difficult to quantify the extent to which Athens
actually protected its tributaries, as our two main sources, Thucydides and Aristotle,
were both quite critical of this imperial period of Athens' history, showing us a
relationship based more on Athens taking resources from the smaller city-states and using
them to fortify itself \footcite[ch. 23--25]{aristotle_constitution}. But we do know that
the Athenians used defending and educating their allies as a rhetorical justification for
their imperial policies, in much the same way as more recent and current empires. When
shown in a positive light, Sophocles' portrayal of Athena acts as an embodiment of
imperial benevolence, lending wisdom and protection to Odysseus and the Greeks.

%% or is she the god of war?

On the other hand, Athena is also the goddess of war. (Or really, one of many gods of war;
I'm not sure there's any Greek god who wasn't at least a little warlike.) In
\textit{Ajax}, Sophocles portrays her as brash, volatile and cruel. These traits are in
tension with her wisdom and her protection of the Greeks. When Odysseus declines to
confront Ajax at the beginning of the play, Athena admonishes him, \textquote{Hold your
peace! Do not earn a reputation for cowardice!} \footcite[l. 75]{ajax} Odysseus appears to
have made a wise and prudent choice, but Athena encourages him, against his judgment, to
expose himself to unnecessary risk with the goal of proving his bravery. When he pushes
back by explaining the risk, she attempts to entice him, \textquote{is not the sweetest
mockery the mockery of enemies?} \footcite[l. 79]{ajax} Athena's reason for encouraging
Odysseus to do something both tactically and strategically foolish is that he will gain
cruel pleasure from Ajax's suffering.

If we imagine Athena as embodying the city, Odysseus as a noble, educated citizen, and
Ajax as a disenfranchised would-be participant, we can understand Athena's violent urges
in this moment as a commentary on Athens' exclusionary dominance over its tributaries
without democratic representation. Athenian foreign policy at the time Sophocles wrote
\textit{Ajax} consisted largely of using the threat of force to convince a nearby
city-state to convert to democratic rule for internal policies, then to extract labor and
resources from it for the nominal purpose of shared defense. Even more than its actual
military might, this policy depends on Athens appearing unfazed by external threats, and
prepared to punish uppity underlings who won't do as they're told. In his portrayal of
Athena, Sophocles stages this Athenian pattern of coercion via the threat of force, the
attitudes it requires of Athens' persona, and the tensions between those attitudes and the
Athenians' stated values.

%% or is she the bringer of rage and ptsd?

%% TRANSITION: relate this to the aforementioned cruelty.
Athena is directly responsible for Ajax's trauma and madness, and
through them for all the conflict in the play. To the modern eye, she seems to be the
goddess of post-traumatic stress and mental illness, with powers over rage, hallucination
and despair. We see her inspiring rage when she tells Odysseus, \textquote{while
  \textins{Ajax} ran about in diseased frenzy, I kept urging him on, kept hurling him into
  the snares of doom} \footcite[ll. 59--60]{ajax}. By her own admission, Athena stoked the
flames of Ajax's rage, driving him to violence and keeping him going after he would
otherwise have stopped. We see her causing hallucinations when she protects the Achaeans
from that same rage, as she \textquote{prevented \textins{Ajax} by casting over his eyes
  oppressive notions of his fatal joy} \footcite[l. 65]{ajax}. Again by her own admission,
Athena shows us that her method of influencing Ajax is to deceive him, to make him see
things that aren't there.

Most sinister of Athena's powers, though, is that Sophocles attributes to her Ajax's
despair and suicidal ideation. Ajax's depressive episode is called \textquote{the anger of
divine Athena} \footcite[l. 757]{ajax} and \textquote{the intolerable anger of the
goddess} \footcite[l. 778]{ajax}, and we are told that her power is great enough to drive
him to suicide within a day. Ajax recognizes that Athena \textquote{abuses me to my
destruction,} and begs for rest and relief from his despair
\footcite[ll. 400--405]{ajax}. This is a chilling depiction of post-traumatic depression,
and Sophocles has quite explicitly attributed it to Athena, goddess of war and
personification of Athens.

%% she's definitely athens, tho

% is war bad?

Looking back through history, we know that Athens' brashness and cruelty led to a
devastating war, but \textit{Ajax} gives us the sense that Sophocles foresaw the same
thing, or at least feared it. At the very least we can say that \textit{Ajax} is an
anti-war play which criticizes Athenian foreign policy for its aggressive expansion, and
we know that Athens' aggressive expansion led to a calamitous war \footcite[ch. 1.23
ll. 1--2]{thucydides_war}. To the effect of warning the Athenians off their disastrous
path, Sophocles stages the horrors of war in \textit{Ajax}. Ajax's own trauma-induced
suicide is most obvious, but not alone. We also see Odysseus shaped by hyperawareness, a
common symptom of PTSD, in the very first line of the play \footcite[l. 1]{ajax}, showing
that all of the Greek soliders are in some way changed by the war, though some less
spectacularly than Ajax. A description of Ajax's rampage makes a brief reference to the
expected treatment of prisoners of war, as Tecmessa tells us, \textquote{Some
\textins{animals} he beheaded; of some he cut the twisted throat or broke the spine;
others he abused in their bonds as though they were men} \footcite[ll. 295--300]{ajax}.
This whole violent image is gut-wrenching, but the worst part is the implication that
abusing captives is normal. And in case all that were not clear enough, the chorus calls
war \textquote{the \textelp{} plague of Ares' detested arms}
\footcite[ll. 1192--1197]{ajax}. Anti-war imagery and messaged pervade \textit{Ajax}.

%% TODO: war will hurt the peasants, like how athena allows ajax to kill herdsmen

%% and does expansionism cause war?

% tell em what you told em
In attributing to Athena Ajax's crimes, illness and death, Sophocles makes a powerful
claim about the outcome of Athens' imperialist, expansionist and militarist actions. He
showed his audience that Athens' violent behaviors were contrary to its stated goals of
democracy and wisdom, and that continuing down that path would lead to a war which would
traumatize both the Athenians and their imperial subjects. As it turns out, he was right,
and the Peloponnesian war devastated Athens and the Hellenic world not long after
\textit{Ajax}.
% EXPLICATE: what claim?
% athens is fucking people up! good people!
% tie in ajax as democratic participant

### 2022-03-18 paper.tex
my final paper for contemp german drama

@@
-% tell em what you're gonna tell em
+Having sat with this question for a bit longer (and reflected on the feedback you gave my
+last essay), and considered the plays I did and didn't like this semester, I think:
+theater because it can be specific where other art forms must be general. Specific in its
+language, specific in its mode, specific in its content. Specific to its place, specific
+to its time, specific to its audience. I'd like to talk about specificity and the
+exploration it allows in four plays we experienced: \textit{Roma Armee}, \textit{The New
+  Gospel}, \textit{Murmel Murmel} and \textit{Black Bismarck}.
 
-In this paper, I will present a reading Sophocles' \textit{Ajax} through the lens of the
-Peloponnesian war. I will justify that lens based on the play's historical context at the
-time it was conceived, and based on Sophocles' life and experiences. I will demonstrate
-that the reading allows us to understand \textit{Ajax}'s treatment of the goddess Athena,
-of militarism and imperialism as a way of life, and of a debate between autocratic and
-democratic constitutions.
+My process in writing this essay was to first select the four plays from our class which
+most affected me, the four about which I still had thoughts I needed to share, and then to
+examine what it was about those plays that affected me. During that process, I settled on
+the realization that the plays which moved me were the ones which spoke to my own lived
+experience, and the plays which frustrated me were those that spoke to an experience I had
+not lived. Having come to that realization, I think I could go back and talk about a
+different set of four plays which might make a stronger argument; I might substitute
+\textit{TANZ} in place of \textit{Murmel Murmel} or \textit{The New Gospel}. But I've
+already written my four reflections, and I've come back to the top to further preface them
+like this.
 
-% why would i think to read Ajax through this lens?
+% I'd like to note, before we begin, that throughout the course of this semester I got my
+% hands on Julia Jarcho's \textit{Writing and the Modern Stage}, and it's shaped the way I
+% think about text, narrative and drama.
 
-%% yes, i know that ajax predates the war.
+Much of my engagement with these plays is going to be in describing the conceptual spaces
+they construct and inhabit. I contend that all art, and in fact all communication, is a
+process of building a conceptual space and populating it with ideas, or from another
+perspective, building a conceptual space which is capable of expressing the ideas you're
+interested in. Math classes do this, computer programs do it, romance novels do it, and
+plays do it. Each of these works builds upon an existing space specific to the person
+consuming it, a space which is informed by expectations and past experiences such as those
+with other works in the same medium or genre. For example, the latest Marvel movie can
+drop quite easily into the space I bring with me because it was defined by the thirty
+Marvel movies I've already seen, whereas a new HBO show must first inform me of the stakes
+and the rules because I don't know whether to expect \textit{Curb Your Enthusiasm} or
+\textit{Westworld}. The unique thing about theater is that a precise understanding of
+one's audience allows a writer or performer to assume more of a shared foundation than is
+possible for artists in other media.
 
-In justifying this lens, I must, most importantly, vie with the fact that \textit{Ajax}
-was almost certainly written before the Peloponnesian war proper. Though dating Sophocles'
-plays is a fraught subject, it is generally accepted that \textit{Ajax} is either the
-earliest or among the earliest of the surviving dramas \footcites[\enquote{Dating the
-Plays}]{meineck_four}[p. 38]{bates_soph}. Recent estimates place the first performance
-of Ajax in the 440s B.C.E. \footcite[p. 10]{finglass_ajax}. At a first glance, this would
-place it well before the Peloponnesian war, which is reliably dated to have begun in 431
-B.C.E. \footcite[p. 32]{lazenby_war}.
+\clearpage{}
 
-My interest is not, however, in the bloody boots-on-the-ground military conflict to which
-Sophocles' later plays respond; rather, in the ongoing tensions between Athenian imperial
-expansion and Sparta's more conservative holdings which sparked the war. Thucydides
-describes a near-constant conflict between the Athenians and the Spartans beginning almost
-immediately after the Persian War \footcite[ch. 1.18]{thucydides_war}. Thucydides, himself
-an Athenian who wrote before, during and shortly after the war, is eager to acknowledge
-that the war was caused by the Spartans' fear of, and desire to contain, Athens' imperial
-expansion  \footcite[ch. 1.23]{thucydides_war}. (An investigation of the social and
-political factors which led Thucydides to publish that claim would be fascinating, but
-beyond the scope of this paper.)
+\paragraph{\textit{Roma Armee}} is both the play that I enjoyed the most out of any we
+watched, and the least interesting among those I'll discuss here. By this I mean, it is
+the most readily recognizable as a play, albeit not a drama, and it is easy to recognize
+both the play's goals and the methods it uses to achieve them. In terms of conceptual
+spaces, \textit{Roma Armee} seemed to occupy the already-extant space of bourgeois
+political theater, and used it to construct some ideas about how to perceive Romani
+people.
 
-%% TODO: what the hell is going on in athens at this time?
+Or at least, that was my experience. Based on our discussion in class, I gather that many
+of my peers did not find \textit{Roma Armee} as straightforward as I do. On closer
+reflection, my specific background prepared me to understand \textit{Roma Armee} more than
+my classmates:
 
-%% FORESHADOW: sophocles foresaw the war, and ajax is his way of warning his peers of the
-%% horrors to come.
+\begin{itemize}
+\item As a practicing reform Jew with meaningful religious education, I have spent long
+  enough studying the Shoah that my understanding goes beyond merely the unexplained
+  deaths of Jews, but includes both knowledge of the propaganda which allowed the Nazis to
+  kill all those people, and an understanding that Jews were not unique, but were the
+  largest among several groups targeted by the Nazis, alongside the Roma. All this to say,
+  I'd been taught, at least vaguely, who the Roma are, and what had been done to them.
 
-%% well, thucydides all but tells me to compare the peloponnesian war w/ the trojan
+\item Growing up, I was into superheroes, especially the X-Men. I was familiar with
+  Magneto's history (and had spent some time critiquing the idea \enquote{what if the
+    Nazi's victims actually were dangerous supernatural beings?}), and I remember being
+  quite upset when the MCU incarnation of Wanda and Pietro Maximoff were generic eastern
+  Europeans instead of Roma.
 
-Thucydides' \textit{History of the Peloponnesian War} also invites this lens by devoting
-much of his first book to comparing the Peloponnesian war with the Trojan
-war \footcite[e.g. ch. 1.23]{thucydides_war}. If we accept that Sophocles was aware of the
-rising tensions between the Athenians and the Spartans when he wrote \textit{Ajax}, then
-it is not difficult to imagine him drawing the same parallel as Thucydides between a
-contemporary conflict and the Greeks' most historicized mythic war. But we need not merely
-imagine, since once one looks through the lens of (pre-) Peloponnesian War Hellenic
-geopolitics, a number of otherwise murky features in the text become clear.
+\item I spent enough time in Sweden (which is basically upstate Germany) to have a sense
+  of how my cultural and historical awareness compared to the norm: especially politically
+  active Europeans would have a deeper understanding of the history than me, as would
+  directly affected minorities, but most had only a passing familiarity with the idea of
+  the Roma.
+\end{itemize}
 
-% demonstrate the lens: Autocracy and democracy
+I believe that \textit{Roma Armee} relied on all of this knowledge, and I also believe
+that, pretty reliably, every person in the Gorki's audience for that performance would
+have had it, although the background that would've given it to them might have been
+different. The superhero thing was probably the least understood among the audience, and
+it's also the thing most explicitly explained by the show.
 
-\textit{Ajax} contains some rather blunt attacks against the coercive nature of autocracy,
-and at least one more subtle reference to the voluntary civic engagement required for
-democracy. The ideological conflict between autocracy and democracy drove the
-Peloponnesian war (or at least, was used to justify it), but categorically could not have
-existed during the Trojan War, as democracy was not invented for several centuries.
+\clearpage{}
 
-%% Menelaus
+\paragraph{\textit{The New Gospel}} doesn't fit my vision for theater. There seemed to be
+some confusion about how to classify it, so let me say: this was a film. Reading Milo
+Rau's manifesto, I can see how he ended up as a filmmaker. The last three of his points
+are devoted to making theater transportable to places other than where it was conceived,
+and to conceiving plays in places other than where they will be performed. The logical
+result of these goals is film.
 
-The two least sympathetic characters in \textit{Ajax}, Menelaus and Agamemnon, both
-espouse (straw men of) autocracy. Menelaus pronounces, \textquote{Now it is, in truth, the
-mark of a base nature when a commoner does not think it right to obey those who stand over
-him. Never can the laws maintain a prosperous course in a city where fear has no fixed
-place} \footcite[ll. 1072--1075]{ajax}. In this moment, Sophocles has a Spartan advocate
-for a rigid social hierarchy without upward mobility, held in place by fear. Menelaus is
-also doing this in the same breath as ordering the desecration of the corpse of a
-traumatized war hero. Spoken before a crowd of Athenian men, most of them
-\enquote{commoners,} and all of them soldiers or veterans, this must have been a powerful
-denouncement of autocracy.
+\begin{enumerate}[label={Command \arabic{*}}, start=8]
+\item \label{raucommand:eight} requires that all plays be transportable in a car or
+  van. Transporting a film requires only a projector and a screen, or even better, a
+  simple digital transfer.
 
-%% Aga 1 - irreverent
+\item requires that plays be conceived in conflict or war zones (or performed there, but
+  that is less interesting, and really just an extension of
+  \ref{raucommand:eight}). \textit{The New Gospel} shows us that Rau's goal is to portray
+  these conflicts to otherwise insulated audiences. Bringing stories from a conflict zone
+  and staging them in Berlin does less to break that insulation than bringing people from
+  the conflict zone and literally placing them onstage in Berlin, which in turn does less
+  than recording images and sounds and recreating them wholesale.
 
-For Agamemnon, not just a Spartan but their king, Sophocles is even more explicit, having
-him say, \textquote{Reverence, I tell you, is not easily practiced by the
-autocrat} \footcite[l. 1350]{ajax}. For a second time in \textit{Ajax}, Sophocles tells
-us that autocratic rule precludes, or at least devalues, just religious observance. In the
-deeply religious society of Hellenic Greece, Sophocles associates autocracy with
-godlessness and wickedness, and thus implicitly joins democracy with piety.
+\item extends \ref{raucommand:eight} by requiring that works actually be transported, not
+  just that they be transportable. And it is here that Rau loses what I love about the
+  theater: his works can no longer laser-target the expectations, experiences and
+  preconceptions of an audience full of Berliners, because they also have to be applicable
+  to the expectations, experiences and preconceptions of Parisians, Luxembourgers and
+  residents of Istanbul (or wherever the Ghent decides to tour).
+\end{enumerate}
 
-%% Aga 2 - shut up & listen
+With \textit{The New Gospel}, Rau set out to break the norms and expectations of theater
+to align with his manifesto, and wound up creating a film. A very beautiful film,
+certainly, but one that lacked the theater's ability to speak to its audience in their own
+language, or to cooperate with its audience in constructing a shared language. We can see
+this, in part, because it was the least referential of all the works I'll discuss: the
+only cultural artifact referenced was the passions, which has got to be the most
+universally known story there is, at least in Europe and the Americas.
 
-In the same moment, Agamemnon also resists the wisdom of Odysseus, telling him,
-\textquote{A good man should listen to those in
-charge} \footcite[l. 1352]{ajax}. Odysseus is not entirely sympathetic in \textit{Ajax},
-but he is still presented as a paragon of wisdom, piety and morality, at least when
-juxtaposed with Agamemnon. And yet Sophocles shows Agamemnon arguing with Odysseus for no
-purpose other than personal vendetta, and with no more of an argument than \enquote{you
-should listen to me because I am king.} This critique of aristocracy, that powerful rulers
-become so rigid and self-assured that they ignore good advice and therefore enact bad
-policies, is typical of both classical and modern democratic rhetoric.
+\clearpage{}
 
-%% ajax - a democratic citizen
+\paragraph{\textit{Murmel Murmel}} makes this list because I can see that it was specific
+to someone's background, but I did not have it. From our discussion, I can see that some
+of my classmates did, though, and I am fascinated by the contrast between my experience
+and theirs. This play seems to have depended on a framework of movement and physicality
+which was present for the dancers and some of the actors in our class, and I suppose for
+people with that framework, the things happening onstage communicated ideas in a way
+analogous to the words and images in \textit{Roma Armee} or \textit{Black Bismarck} did to
+me. For me, though, \textit{Murmel Murmel} was like reading a book in a foreign
+language. Or maybe like being given the instruction manual for an IKEA shelf but having to
+buy the parts from a Home Depot, and what you end up with just isn't right.
 
-Sophocles presents Ajax's involvement in the Trojan war as the willing participation of an
-outsider in democracy. When defending Ajax against Menelaus' accusations of
-insubordination, Teucer says, \textquote{Did he not sail of his own accord? As his own
-master? On what grounds are you his commander? \textelp{} Nowhere was it established among
-your lawful powers that you should order him any more than he
-you} \footcite[ll. 1099--1104]{ajax}. This defense of personal sovereignty, that each
-citizen/soldier chose to participate in and contribute to society/the war effort freely,
-without duress, and without binding themselves to the will of an individual, could have
-come from the textbook of my freshman gen ed course on mixed constitutions. And to
-Sophocles, this is what makes Ajax noble: he makes his choices for no reason except that
-he thinks they are right, and he collaborates with others to achieve larger goals, even
-when personal or ideological differences lead to disagreement.
-%% CONSIDER: tie this back in to ``all this doesn't make sense in the actual trojan war,
-%% bc a. democracy didn't exist yet, and b. thucydides tells us that agamemnon totally did
-%% coerce all his allies''
+For the purposes of this assignment, I suppose I can let \textit{Murmel Murmel} serve as a
+counterpoint, an exemplar of the downside of the hyperspecificity I'm describing. When a
+play is for a specific group of people, and it depends on those peoples' foreknowledge,
+any outsiders will find it more or less unintelligible. In addition to being better able
+to predict its audience relative to film or other media, the theater is also, in my
+opinion, more \enquote{allowed,} culturally, to be exclusionary in this way.  I think back
+to some discussions I had about the film \textit{The Green Knight}. I loved the movie, but
+I think I was only able to love it because I was used to post-dramatic theater, where plot
+is not the guiding force. My friends and family who lacked that background did not like
+\textit{The Green Knight}, and moreover, they seemed angry that there were movies in
+theaters which weren't for them, which depended on a conceptual framework they didn't
+have. For all I complain about \textit{Murmel Murmel}, I was certainly not surprised that
+I didn't get it; that's just par for the course in the theater.
 
-The ideological conflict between autocracy and democracy justifies this lens, in that the
-Peloponnesian war explains easily-recognizable features of the text which make little
-sense without the political context of Sophocles' time, but it is not terribly exciting to
-learn that a Sophocles play promotes democracy and decries autocracy. To learn anything
-new, we must focus our lens on Sophocles' treatment of Athenian militarism and
-imperialism, and of the city's god, Athena.
+\clearpage{}
 
-% gain insight from the lens:
-% athena - what's her deal?
+\paragraph{\textit{Black Bismarck},} more than anything else we've seen in this course, is
+my vision for post-dramatic theater. I was not within its crosshairs, nor was I quite as
+close as I was for \textit{Roma Armee}, but I was certainly much closer than I was for
+\textit{Murmel Murmel} or \textit{TANZ}. My language requirement in German left me at
+least aware of the linguistic and cultural phenomena the play was responding to.  I
+imagine the actual audience, when shown the slideshow of all the monuments to Bismarck,
+would have had a realization like, \enquote{huh, I walk past these all the time without
+  ever looking too closely or questioning why they were there. I guess I never noticed
+  just how prevalent Bismarck monuments are,} but since COVID spoiled my study abroad
+plans, I never actually walked past any of the statues. I had seen \textit{Ghostbusters},
+but I had to be reminded of the specific context of Stay-Puft's appearance; I had
+forgotten about \enquote{clear your mind!}
 
-%% is she the noble goddess of wisdom?
+But after investigating each of these things, I think I can imagine how \textit{Black
+  Bismarck} affected its audience, and I am impressed. It seems likely to me that much of
+the audience had traveled via that actual subway station, or at the very least one like
+it, and had seen the statues that the white rabbit kicked over. A film could not have been
+sure of that; filmmakers must contend with the possibility that the same version of
+products will be shown in theaters across the country or the world, to people of wildly
+different backgrounds, and even in informal settings like living rooms and
+airplanes. (Ironically, that's exactly our experience with \textit{Black Bismarck}.)
 
-\textit{Ajax} is absolutely rife with references to Athena, who seems to play two somewhat
-distinct roles. On the one hand, she is the patron of Odysseus and defender of the Danaans
-(who, through this lens, must be taken to represent the Athenians in Sophocles'
-time). Odysseus tells us that Athena is \textquote{dearest to \textins{him} of the
-  gods} \footcite[l. 14]{ajax}, and later in the play, the chorus praises him for his
-wisdom \footcite[ll. 1374--1375]{ajax}, a trait typically associated with Athena. His
-Athenian wisdom allows Odysseus to see past the hatred which has blinded Menelaus and
-Agamemnon, and to resolve the drama of \textit{Ajax} in an honorable, productive
-way.
+But (when performed live), \textit{Black Bismarck} could assume those things, and it did,
+to great effect. To an actual Berliner, who spoke German as a relatively uncritical
+first-language speaker, who experienced the city and its monuments as the background of
+everyday life, this play must have masterfully invited (or forced) them to step back and
+to critically reexamine their world.
 
-Athena also tells us that it was her intervention which prevented Ajax from slaughtering
-the Greek army, reducing the impact of his crimes to a handful of sheep and herdsmen
-\footcite[ll. 40--75]{ajax}. If we read Athena as a personification of the city, we get a
-picture of Athens as the imperial protector, using its wisdom to avert conflicts and
-catastrophes among its subjects. It's difficult to quantify the extent to which Athens
-actually protected its tributaries, as our two main sources, Thucydides and Aristotle,
-were both quite critical of this imperial period of Athens' history, showing us a
-relationship based more on Athens taking resources from the smaller city-states and using
-them to fortify itself \footcite[ch. 23--25]{aristotle_constitution}. But we do know that
-the Athenians used defending and educating their allies as a rhetorical justification for
-their imperial policies, in much the same way as more recent and current empires. When
-shown in a positive light, Sophocles' portrayal of Athena acts as an embodiment of
-imperial benevolence, lending wisdom and protection to Odysseus and the Greeks.
+\clearpage{}
 
-%% or is she the god of war?
-
-On the other hand, Athena is also the goddess of war. (Or really, one of many gods of war;
-I'm not sure there's any Greek god who wasn't at least a little warlike.) In
-\textit{Ajax}, Sophocles portrays her as brash, volatile and cruel. These traits are in
-tension with her wisdom and her protection of the Greeks. When Odysseus declines to
-confront Ajax at the beginning of the play, Athena admonishes him, \textquote{Hold your
-peace! Do not earn a reputation for cowardice!} \footcite[l. 75]{ajax} Odysseus appears to
-have made a wise and prudent choice, but Athena encourages him, against his judgment, to
-expose himself to unnecessary risk with the goal of proving his bravery. When he pushes
-back by explaining the risk, she attempts to entice him, \textquote{is not the sweetest
-mockery the mockery of enemies?} \footcite[l. 79]{ajax} Athena's reason for encouraging
-Odysseus to do something both tactically and strategically foolish is that he will gain
-cruel pleasure from Ajax's suffering.
-
-If we imagine Athena as embodying the city, Odysseus as a noble, educated citizen, and
-Ajax as a disenfranchised would-be participant, we can understand Athena's violent urges
-in this moment as a commentary on Athens' exclusionary dominance over its tributaries
-without democratic representation. Athenian foreign policy at the time Sophocles wrote
-\textit{Ajax} consisted largely of using the threat of force to convince a nearby
-city-state to convert to democratic rule for internal policies, then to extract labor and
-resources from it for the nominal purpose of shared defense. Even more than its actual
-military might, this policy depends on Athens appearing unfazed by external threats, and
-prepared to punish uppity underlings who won't do as they're told. In his portrayal of
-Athena, Sophocles stages this Athenian pattern of coercion via the threat of force, the
-attitudes it requires of Athens' persona, and the tensions between those attitudes and the
-Athenians' stated values.
-
-%% or is she the bringer of rage and ptsd?
-
-%% TRANSITION: relate this to the aforementioned cruelty.
-Athena is directly responsible for Ajax's trauma and madness, and
-through them for all the conflict in the play. To the modern eye, she seems to be the
-goddess of post-traumatic stress and mental illness, with powers over rage, hallucination
-and despair. We see her inspiring rage when she tells Odysseus, \textquote{while
-  \textins{Ajax} ran about in diseased frenzy, I kept urging him on, kept hurling him into
-  the snares of doom} \footcite[ll. 59--60]{ajax}. By her own admission, Athena stoked the
-flames of Ajax's rage, driving him to violence and keeping him going after he would
-otherwise have stopped. We see her causing hallucinations when she protects the Achaeans
-from that same rage, as she \textquote{prevented \textins{Ajax} by casting over his eyes
-  oppressive notions of his fatal joy} \footcite[l. 65]{ajax}. Again by her own admission,
-Athena shows us that her method of influencing Ajax is to deceive him, to make him see
-things that aren't there.
-
-Most sinister of Athena's powers, though, is that Sophocles attributes to her Ajax's
-despair and suicidal ideation. Ajax's depressive episode is called \textquote{the anger of
-divine Athena} \footcite[l. 757]{ajax} and \textquote{the intolerable anger of the
-goddess} \footcite[l. 778]{ajax}, and we are told that her power is great enough to drive
-him to suicide within a day. Ajax recognizes that Athena \textquote{abuses me to my
-destruction,} and begs for rest and relief from his despair
-\footcite[ll. 400--405]{ajax}. This is a chilling depiction of post-traumatic depression,
-and Sophocles has quite explicitly attributed it to Athena, goddess of war and
-personification of Athens.
-
-%% she's definitely athens, tho
-
-% is war bad?
-
-Looking back through history, we know that Athens' brashness and cruelty led to a
-devastating war, but \textit{Ajax} gives us the sense that Sophocles foresaw the same
-thing, or at least feared it. At the very least we can say that \textit{Ajax} is an
-anti-war play which criticizes Athenian foreign policy for its aggressive expansion, and
-we know that Athens' aggressive expansion led to a calamitous war \footcite[ch. 1.23
-ll. 1--2]{thucydides_war}. To the effect of warning the Athenians off their disastrous
-path, Sophocles stages the horrors of war in \textit{Ajax}. Ajax's own trauma-induced
-suicide is most obvious, but not alone. We also see Odysseus shaped by hyperawareness, a
-common symptom of PTSD, in the very first line of the play \footcite[l. 1]{ajax}, showing
-that all of the Greek soliders are in some way changed by the war, though some less
-spectacularly than Ajax. A description of Ajax's rampage makes a brief reference to the
-expected treatment of prisoners of war, as Tecmessa tells us, \textquote{Some
-\textins{animals} he beheaded; of some he cut the twisted throat or broke the spine;
-others he abused in their bonds as though they were men} \footcite[ll. 295--300]{ajax}.
-This whole violent image is gut-wrenching, but the worst part is the implication that
-abusing captives is normal. And in case all that were not clear enough, the chorus calls
-war \textquote{the \textelp{} plague of Ares' detested arms}
-\footcite[ll. 1192--1197]{ajax}. Anti-war imagery and messaged pervade \textit{Ajax}.
-
-%% TODO: war will hurt the peasants, like how athena allows ajax to kill herdsmen
-
-%% and does expansionism cause war?
-
-% tell em what you told em
-In attributing to Athena Ajax's crimes, illness and death, Sophocles makes a powerful
-claim about the outcome of Athens' imperialist, expansionist and militarist actions. He
-showed his audience that Athens' violent behaviors were contrary to its stated goals of
-democracy and wisdom, and that continuing down that path would lead to a war which would
-traumatize both the Athenians and their imperial subjects. As it turns out, he was right,
-and the Peloponnesian war devastated Athens and the Hellenic world not long after
-\textit{Ajax}.
-% EXPLICATE: what claim?
-% athens is fucking people up! good people!
-% tie in ajax as democratic participant
+For my money, the advantage of theater's specificity is that plays have to do less work
+establishing a shared context before they can get to something new and interesting. They
+can be a scalpel instead of a chainsaw. If that's not enough for you, then in this space,
+feel free to imagine a conclusion about how great theater is which uses the phrases
+\enquote{refreshing contrast,} \enquote{intimate,} \enquote{mass-market entertainment,}
+\enquote{globalization,} and \enquote{more important than ever.}

### 2022-03-18 paper.tex
a paper i wrote a while ago

@@
-Having sat with this question for a bit longer (and reflected on the feedback you gave my
-last essay), and considered the plays I did and didn't like this semester, I think:
-theater because it can be specific where other art forms must be general. Specific in its
-language, specific in its mode, specific in its content. Specific to its place, specific
-to its time, specific to its audience. I'd like to talk about specificity and the
-exploration it allows in four plays we experienced: \textit{Roma Armee}, \textit{The New
-  Gospel}, \textit{Murmel Murmel} and \textit{Black Bismarck}.
+What I propose is not, strictly, a unit of study, but rather, a focus that I would like to
+see incorporated into the units already in your class. I find myself fascinated by the
+economic realities of performance art, and the ways different systems of funding affect
+the works produced. I was struck by this realization during a class from Brandon Woolf and
+Andy Goldberg a few weeks ago, when we discussed Berlin's \textit{Freie Szene}. The norm
+in Germany is for each theater to permanently employ an ensemble, who are paid a stable
+salary by the state. The \textit{Freie Szene}, in contrast, consists of independent
+troupes who are funded by government grants on a per-show basis. On the one hand,
+\textit{freie} groups exist in a much more precarious space, where a series of rejected
+proposals can leave them missing rent or collecting welfare. On the other hand, they enjoy
+significant freedoms (hence the name) because of their disconnect from the institutional
+hierarchies which employ more traditional ensembles. Notably, it is challenging for a
+director working with a salaried ensemble to justify bringing in outside collaborators, so
+temporarily hiring subject matter experts to collaborate on particular projects has become
+a hallmark of the \textit{Freie Szene}. We got to see this play out onstage: \textit{Roma
+  Armee}, a play interested in staging Romani performers, wound up with a cast that was
+roughly half Romani, because the Gorki's ensemble only employed a handful of Romani
+people. Rimini Protokoll's \textit{100\% Berlin}, on the other hand, invited and placed
+onstage 100 Berlin residents. (In truth, I'm not sure if they were paid or volunteers, but
+still.) I am interested in applying this same lens to the cultures of performance art we
+discussed through the semester.
 
-My process in writing this essay was to first select the four plays from our class which
-most affected me, the four about which I still had thoughts I needed to share, and then to
-examine what it was about those plays that affected me. During that process, I settled on
-the realization that the plays which moved me were the ones which spoke to my own lived
-experience, and the plays which frustrated me were those that spoke to an experience I had
-not lived. Having come to that realization, I think I could go back and talk about a
-different set of four plays which might make a stronger argument; I might substitute
-\textit{TANZ} in place of \textit{Murmel Murmel} or \textit{The New Gospel}. But I've
-already written my four reflections, and I've come back to the top to further preface them
-like this.
+Throughout this essay, I will perform a relatively shallow investigation of various
+economies of performance. I will compare these economies and their results to the
+performance economy most familiar to me, Broadway (and American drama more broadly, as I
+grew up in Minneapolis), but note that this is merely an artifact of convenience. My goal
+is not to privilege Broadway as the optimal performance economy to which all history has
+led; in fact, I think it's actually a pretty terrible version, and I find many of the
+economies I'll discuss much more compelling.
 
-% I'd like to note, before we begin, that throughout the course of this semester I got my
-% hands on Julia Jarcho's \textit{Writing and the Modern Stage}, and it's shaped the way I
-% think about text, narrative and drama.
+Also note that most of what I write here is my own analysis based on my background
+knowledge and what we discussed in class, along with a non-zero but frankly insufficient
+amount of research. I caution you that what I write below will likely not reflect
+scholarly consensus, because I'm not sure what the scholarly consensus is.\footnote{I
+  blame this at least partially on the focus on novelty and reinterpretation in the
+  humanities, which the computer scientist in me is unused to. In my discipline, it's not
+  uncommon for the academic community to come to an agreement about the right way to do
+  something, and then to stop worrying about it and move on to other things. This does not
+  appear to be the case in theater studies or classics.}
 
-Much of my engagement with these plays is going to be in describing the conceptual spaces
-they construct and inhabit. I contend that all art, and in fact all communication, is a
-process of building a conceptual space and populating it with ideas, or from another
-perspective, building a conceptual space which is capable of expressing the ideas you're
-interested in. Math classes do this, computer programs do it, romance novels do it, and
-plays do it. Each of these works builds upon an existing space specific to the person
-consuming it, a space which is informed by expectations and past experiences such as those
-with other works in the same medium or genre. For example, the latest Marvel movie can
-drop quite easily into the space I bring with me because it was defined by the thirty
-Marvel movies I've already seen, whereas a new HBO show must first inform me of the stakes
-and the rules because I don't know whether to expect \textit{Curb Your Enthusiasm} or
-\textit{Westworld}. The unique thing about theater is that a precise understanding of
-one's audience allows a writer or performer to assume more of a shared foundation than is
-possible for artists in other media.
+\section*{5th century B.C.E. Athens}
+% case study: greek drama
 
-\clearpage{}
+%% theater as a public service, performed by soldiers
 
-\paragraph{\textit{Roma Armee}} is both the play that I enjoyed the most out of any we
-watched, and the least interesting among those I'll discuss here. By this I mean, it is
-the most readily recognizable as a play, albeit not a drama, and it is easy to recognize
-both the play's goals and the methods it uses to achieve them. In terms of conceptual
-spaces, \textit{Roma Armee} seemed to occupy the already-extant space of bourgeois
-political theater, and used it to construct some ideas about how to perceive Romani
-people.
+Our class, like every theater history class, started with the Greeks, and so I suppose I
+should, too. My understanding, based largely on a course I took with Peter Meineck to
+satisfy my pre-1900s theater requirement, is that Athenian drama, was written by private
+citizens for prestige, and performed by enlisted soldiers as part of their duties to the
+city. I've also heard it proposed that wealthy Athenians would fund the playwrights, but
+this was again with the aim of social prestige, not in exchange for artistic control or
+seeking some return on investment. It's difficult to make broad statements about the
+effects of this cultural organization on its products, as we have access to only a small
+subset of the plays written by the three most successful playwrights, and none by any of
+their myriad competitors. Still, we can trace some through-lines.
 
-Or at least, that was my experience. Based on our discussion in class, I gather that many
-of my peers did not find \textit{Roma Armee} as straightforward as I do. On closer
-reflection, my specific background prepared me to understand \textit{Roma Armee} more than
-my classmates:
+Athenian dramas tended to be quite specific to the year they were written and staged; I
+think of \textit{Ajax} commenting on the rising tensions leading up to the Peloponnesian
+war, and of \textit{Oedipus Tyrannus} responding to Pericles' death in the Plague of
+Athens. Hyper-specific plays like this are discouraged in many other cultural economies,
+including our own, by a desire that scripts be reusable and accessible. A Broadway show
+which spoke deeply to its opening-night audience, but within a year started to alienate
+all newcomers would be considered a failure, but the Athenian model allowed plays to do
+just that.
 
-\begin{itemize}
-\item As a practicing reform Jew with meaningful religious education, I have spent long
-  enough studying the Shoah that my understanding goes beyond merely the unexplained
-  deaths of Jews, but includes both knowledge of the propaganda which allowed the Nazis to
-  kill all those people, and an understanding that Jews were not unique, but were the
-  largest among several groups targeted by the Nazis, alongside the Roma. All this to say,
-  I'd been taught, at least vaguely, who the Roma are, and what had been done to them.
+Being performed as a public service, without a direct financial incentive like ticket
+sales, seems to have encouraged Athenian playwrights to focus more on exposing their
+audiences to new ideas or perspectives, and less on leaving them entertained or happy,
+than the Broadway scene. I think once again of \textit{Ajax}, this time the ambiguous
+light in which it portrays Ajax, one of Athens' founding heroes; Odysseus, arguably the
+most beloved mythical hero; and Athena, the patron goddess of Athens. I'd love to tell you
+what the Greek public thought of \textit{Ajax}, but all I can tell you is that there is no
+record of its having won the City Dionysia. We do know that some very challenging and
+provocative plays won first place, including Aeschylus' \textit{Persians}, which invited
+sympathy for a hated enemy against whom Athens had quite recently fought and won a brutal
+war. This suggests to me that the festival's convoluted judging scheme managed to select
+for some definition of didacticism and artistic integrity, rather than pure entertainment
+value.
 
-\item Growing up, I was into superheroes, especially the X-Men. I was familiar with
-  Magneto's history (and had spent some time critiquing the idea \enquote{what if the
-    Nazi's victims actually were dangerous supernatural beings?}), and I remember being
-  quite upset when the MCU incarnation of Wanda and Pietro Maximoff were generic eastern
-  Europeans instead of Roma.
+\section*{Imperial Rome}
 
-\item I spent enough time in Sweden (which is basically upstate Germany) to have a sense
-  of how my cultural and historical awareness compared to the norm: especially politically
-  active Europeans would have a deeper understanding of the history than me, as would
-  directly affected minorities, but most had only a passing familiarity with the idea of
-  the Roma.
-\end{itemize}
+Roman theater, as I understand it, is best summarized by the phrase \enquote{bread and
+  circuses.} Roman plays were funded either by the empire itself or by the patricians, and
+served (with a range of conscious forethought) to placate and distract the plebeians and
+to maintain the social hierarchy. At least as the post-Roman Christians tell it, this
+meant the Romans' theater was gaudy, excessive and meaningless. Their main contribution to
+drama as an art was in importing, preserving and re-performing the works of other
+cultures; every surviving Greek play comes to us through the Romans. If I wanted to be
+particularly political, I might set aside some time to draw parallels with modern America,
+with the message that empire does not promote compelling art. That essay sounds like a
+bummer, though, so instead I'll move on.
 
-I believe that \textit{Roma Armee} relied on all of this knowledge, and I also believe
-that, pretty reliably, every person in the Gorki's audience for that performance would
-have had it, although the background that would've given it to them might have been
-different. The superhero thing was probably the least understood among the audience, and
-it's also the thing most explicitly explained by the show.
+\section*{Medieval religious drama}
 
-\clearpage{}
+Frankly, I'm not that interested in this era, which is just as well because there's very
+little to say about it. Medieval European history is characterized by the Church's
+overwhelming influence over art, culture and day-to-day life. The Romans' excesses led the
+Church to reject theater for the better part of a millennium, and when the art does
+re-emerge, we see morality plays like \textit{Everyman}. It seems we actually know very
+little about \textit{Everyman}; we don't know who wrote it, who first performed it, where
+it was performed, or who paid for any of these things. I do feel safe to assert, though,
+that the church was directly involved, either in a positive role by sponsoring the play's
+creation or a negative one in restricting its content. Comparing this period with Broadway
+theater is difficult, as our society is so secular compared to medieval Europe.
 
-\paragraph{\textit{The New Gospel}} doesn't fit my vision for theater. There seemed to be
-some confusion about how to classify it, so let me say: this was a film. Reading Milo
-Rau's manifesto, I can see how he ended up as a filmmaker. The last three of his points
-are devoted to making theater transportable to places other than where it was conceived,
-and to conceiving plays in places other than where they will be performed. The logical
-result of these goals is film.
+\section*{Shakespeare and the renaissance}
 
-\begin{enumerate}[label={Command \arabic{*}}, start=8]
-\item \label{raucommand:eight} requires that all plays be transportable in a car or
-  van. Transporting a film requires only a projector and a screen, or even better, a
-  simple digital transfer.
+It's in the renaissance that we first see a theater economy similar to the one I live with
+today, and by Shakespeare's time in the Elizabethan age, the parallels are
+remarkable. Playwrights were commissioned by wealthy patrons, especially the monarchs, and
+used that funding to support initial performances. They earned profit from ticket sales,
+and used the prestige of popular shows to attract patrons for future works. The notable
+difference compared to Broadway is that Shakespeare had no obligation to pay back his
+patrons' gifts; they were not investors. But the model is fundamentally the same:
+monarchs, aristocrats and other people of wealth used financial involvement to dictate the
+content of new plays, and ticket sales provided an incentive to produce popular,
+crowd-pleasing shows.
 
-\item requires that plays be conceived in conflict or war zones (or performed there, but
-  that is less interesting, and really just an extension of
-  \ref{raucommand:eight}). \textit{The New Gospel} shows us that Rau's goal is to portray
-  these conflicts to otherwise insulated audiences. Bringing stories from a conflict zone
-  and staging them in Berlin does less to break that insulation than bringing people from
-  the conflict zone and literally placing them onstage in Berlin, which in turn does less
-  than recording images and sounds and recreating them wholesale.
+Patrons in this era used drama as a means of political conflict and propaganda. On the
+world stage, this takes the form of the monarchs commissioning plays which glorify,
+historicize and mythologize England's past. We can't actually say for certain which of
+Shakespeare's plays were commissioned by Elizabeth or James, or at least, I can't find a
+source which justifies the claim, but we do know that a disproportionate number of
+Shakespeare's plays are devoted to England's frankly very boring history, and that the
+monarchs were his primary patrons. More interesting, on a local level, we see at least one
+instance of an internal dispute using a Shakespeare play to justify an attempted coup: the
+Earl of Essex famously hired Shakespeare's company, the Chamberlain's Men, to perform
+\textit{Richard II} in the hopes of swaying public opinion towards the would-be usurper
+against Elizabeth. (It didn't work.)
 
-\item extends \ref{raucommand:eight} by requiring that works actually be transported, not
-  just that they be transportable. And it is here that Rau loses what I love about the
-  theater: his works can no longer laser-target the expectations, experiences and
-  preconceptions of an audience full of Berliners, because they also have to be applicable
-  to the expectations, experiences and preconceptions of Parisians, Luxembourgers and
-  residents of Istanbul (or wherever the Ghent decides to tour).
-\end{enumerate}
-
-With \textit{The New Gospel}, Rau set out to break the norms and expectations of theater
-to align with his manifesto, and wound up creating a film. A very beautiful film,
-certainly, but one that lacked the theater's ability to speak to its audience in their own
-language, or to cooperate with its audience in constructing a shared language. We can see
-this, in part, because it was the least referential of all the works I'll discuss: the
-only cultural artifact referenced was the passions, which has got to be the most
-universally known story there is, at least in Europe and the Americas.
-
-\clearpage{}
-
-\paragraph{\textit{Murmel Murmel}} makes this list because I can see that it was specific
-to someone's background, but I did not have it. From our discussion, I can see that some
-of my classmates did, though, and I am fascinated by the contrast between my experience
-and theirs. This play seems to have depended on a framework of movement and physicality
-which was present for the dancers and some of the actors in our class, and I suppose for
-people with that framework, the things happening onstage communicated ideas in a way
-analogous to the words and images in \textit{Roma Armee} or \textit{Black Bismarck} did to
-me. For me, though, \textit{Murmel Murmel} was like reading a book in a foreign
-language. Or maybe like being given the instruction manual for an IKEA shelf but having to
-buy the parts from a Home Depot, and what you end up with just isn't right.
-
-For the purposes of this assignment, I suppose I can let \textit{Murmel Murmel} serve as a
-counterpoint, an exemplar of the downside of the hyperspecificity I'm describing. When a
-play is for a specific group of people, and it depends on those peoples' foreknowledge,
-any outsiders will find it more or less unintelligible. In addition to being better able
-to predict its audience relative to film or other media, the theater is also, in my
-opinion, more \enquote{allowed,} culturally, to be exclusionary in this way.  I think back
-to some discussions I had about the film \textit{The Green Knight}. I loved the movie, but
-I think I was only able to love it because I was used to post-dramatic theater, where plot
-is not the guiding force. My friends and family who lacked that background did not like
-\textit{The Green Knight}, and moreover, they seemed angry that there were movies in
-theaters which weren't for them, which depended on a conceptual framework they didn't
-have. For all I complain about \textit{Murmel Murmel}, I was certainly not surprised that
-I didn't get it; that's just par for the course in the theater.
-
-\clearpage{}
-
-\paragraph{\textit{Black Bismarck},} more than anything else we've seen in this course, is
-my vision for post-dramatic theater. I was not within its crosshairs, nor was I quite as
-close as I was for \textit{Roma Armee}, but I was certainly much closer than I was for
-\textit{Murmel Murmel} or \textit{TANZ}. My language requirement in German left me at
-least aware of the linguistic and cultural phenomena the play was responding to.  I
-imagine the actual audience, when shown the slideshow of all the monuments to Bismarck,
-would have had a realization like, \enquote{huh, I walk past these all the time without
-  ever looking too closely or questioning why they were there. I guess I never noticed
-  just how prevalent Bismarck monuments are,} but since COVID spoiled my study abroad
-plans, I never actually walked past any of the statues. I had seen \textit{Ghostbusters},
-but I had to be reminded of the specific context of Stay-Puft's appearance; I had
-forgotten about \enquote{clear your mind!}
-
-But after investigating each of these things, I think I can imagine how \textit{Black
-  Bismarck} affected its audience, and I am impressed. It seems likely to me that much of
-the audience had traveled via that actual subway station, or at the very least one like
-it, and had seen the statues that the white rabbit kicked over. A film could not have been
-sure of that; filmmakers must contend with the possibility that the same version of
-products will be shown in theaters across the country or the world, to people of wildly
-different backgrounds, and even in informal settings like living rooms and
-airplanes. (Ironically, that's exactly our experience with \textit{Black Bismarck}.)
-
-But (when performed live), \textit{Black Bismarck} could assume those things, and it did,
-to great effect. To an actual Berliner, who spoke German as a relatively uncritical
-first-language speaker, who experienced the city and its monuments as the background of
-everyday life, this play must have masterfully invited (or forced) them to step back and
-to critically reexamine their world.
-
-\clearpage{}
-
-For my money, the advantage of theater's specificity is that plays have to do less work
-establishing a shared context before they can get to something new and interesting. They
-can be a scalpel instead of a chainsaw. If that's not enough for you, then in this space,
-feel free to imagine a conclusion about how great theater is which uses the phrases
-\enquote{refreshing contrast,} \enquote{intimate,} \enquote{mass-market entertainment,}
-\enquote{globalization,} and \enquote{more important than ever.}
+The popular incentive of ticket sales mostly shows through in Shakespeare's dirty jokes
+and reliance on convention. You've already read my take on the dirty jokes, though I'll
+point out that they show up not only in \textit{Romeo and Juliet} but also in otherwise
+serious tragedies; \textit{Theater Histories} cites the porter scene in
+\textit{Macbeth}. In addition to saucy entertainment, audiences expected certain
+conventions, especially in terms of plot and style. Comedies had to end with a marriage,
+and tragedies with a death; \textit{Theater Histories} here points out that \textit{As You
+  Like It} and \textit{Hamlet} had four each of the appropriate ending. Jeremy Lopez has a
+whole book about the conventions expected by theater audiences at this time, and the way
+playwrights (especially Shakespeare) sought to meet those expectations; he highlights puns
+and wordplay, asides and soliloquy, and rapid tonal shifts to keep things exciting. Other
+than the specific plot conventions, which have changed to accommodate a new zeitgeist, I
+think this whole paragraph could apply to Broadway musicals (dramas tend to be more
+experimental and less constrained); they're filled with crowd-pleasing plot moments like
+the long-awaited kiss at the end, dirty jokes, songs which illustrate a character's
+internal monologue, and pretty much every other device Lopez describes.

### 2022-03-18 paper.tex
a paper i wrote a while ago

@@
-What I propose is not, strictly, a unit of study, but rather, a focus that I would like to
-see incorporated into the units already in your class. I find myself fascinated by the
-economic realities of performance art, and the ways different systems of funding affect
-the works produced. I was struck by this realization during a class from Brandon Woolf and
-Andy Goldberg a few weeks ago, when we discussed Berlin's \textit{Freie Szene}. The norm
-in Germany is for each theater to permanently employ an ensemble, who are paid a stable
-salary by the state. The \textit{Freie Szene}, in contrast, consists of independent
-troupes who are funded by government grants on a per-show basis. On the one hand,
-\textit{freie} groups exist in a much more precarious space, where a series of rejected
-proposals can leave them missing rent or collecting welfare. On the other hand, they enjoy
-significant freedoms (hence the name) because of their disconnect from the institutional
-hierarchies which employ more traditional ensembles. Notably, it is challenging for a
-director working with a salaried ensemble to justify bringing in outside collaborators, so
-temporarily hiring subject matter experts to collaborate on particular projects has become
-a hallmark of the \textit{Freie Szene}. We got to see this play out onstage: \textit{Roma
-  Armee}, a play interested in staging Romani performers, wound up with a cast that was
-roughly half Romani, because the Gorki's ensemble only employed a handful of Romani
-people. Rimini Protokoll's \textit{100\% Berlin}, on the other hand, invited and placed
-onstage 100 Berlin residents. (In truth, I'm not sure if they were paid or volunteers, but
-still.) I am interested in applying this same lens to the cultures of performance art we
-discussed through the semester.
+\newcommand{\theplay}{\textit{Romeo and Juliet}}
+\newcommand{\billy}{Shakespeare}
+\newcommand{\theproduction}{\textit{Slapstick Shakespeare}}
 
-Throughout this essay, I will perform a relatively shallow investigation of various
-economies of performance. I will compare these economies and their results to the
-performance economy most familiar to me, Broadway (and American drama more broadly, as I
-grew up in Minneapolis), but note that this is merely an artifact of convenience. My goal
-is not to privilege Broadway as the optimal performance economy to which all history has
-led; in fact, I think it's actually a pretty terrible version, and I find many of the
-economies I'll discuss much more compelling.
+I read \theplay{} in high school, led by an English teacher who frankly wasn't very
+good. She taught the play as a serious tragic romance. I experienced it more as a dark
+comedy; an elaborate farce at the expense of young people. I was aided by a good annotated
+edition (I've unfortunately been unable to find the same one since), which explained all
+the dirty jokes.
 
-Also note that most of what I write here is my own analysis based on my background
-knowledge and what we discussed in class, along with a non-zero but frankly insufficient
-amount of research. I caution you that what I write below will likely not reflect
-scholarly consensus, because I'm not sure what the scholarly consensus is.\footnote{I
-  blame this at least partially on the focus on novelty and reinterpretation in the
-  humanities, which the computer scientist in me is unused to. In my discipline, it's not
-  uncommon for the academic community to come to an agreement about the right way to do
-  something, and then to stop worrying about it and move on to other things. This does not
-  appear to be the case in theater studies or classics.}
+The plot of \theplay{} always struck me as being... less than romantic. The supposed
+\enquote{star-crossed lovers} are a pair of very young teenagers (Juliet is 13) who become
+instantly infatuated upon meeting at a party. In fact, it's not even the first time Romeo
+has fallen hopelessly in love! I'd like to stage the play in such a way to highlight the
+absurdity, bring the raunchy jokes to a forefront, and hopefully destroy the canonical
+reading that it's a sincere romance.
 
-\section*{5th century B.C.E. Athens}
-% case study: greek drama
+The dramatic literature major in me feels obligated to describe my approach as a comedic
+\textit{Verfremdungseffekt}. Similar to how Brechtian drama imposes distance in order to
+force its audience to critically reconsider the social and political circumstances
+presented, \theproduction{} will force its audience to critically reconsider the canonical
+interpretation of \theplay{}, particularly in terms of genre and tone.
 
-%% theater as a public service, performed by soldiers
+\section*{A new prologue}
 
-Our class, like every theater history class, started with the Greeks, and so I suppose I
-should, too. My understanding, based largely on a course I took with Peter Meineck to
-satisfy my pre-1900s theater requirement, is that Athenian drama, was written by private
-citizens for prestige, and performed by enlisted soldiers as part of their duties to the
-city. I've also heard it proposed that wealthy Athenians would fund the playwrights, but
-this was again with the aim of social prestige, not in exchange for artistic control or
-seeking some return on investment. It's difficult to make broad statements about the
-effects of this cultural organization on its products, as we have access to only a small
-subset of the plays written by the three most successful playwrights, and none by any of
-their myriad competitors. Still, we can trace some through-lines.
+Rather than a chorus, I think I'll read the two prologues (there's one before each of the
+first two acts), wearing a mobcap with a feather, to signify my being in-character as the
+Bard. Before the first prologue, I'll enter holding the hat, to signify being
+out-of-character, and address the audience with an introduction to set the tone:
 
-Athenian dramas tended to be quite specific to the year they were written and staged; I
-think of \textit{Ajax} commenting on the rising tensions leading up to the Peloponnesian
-war, and of \textit{Oedipus Tyrannus} responding to Pericles' death in the Plague of
-Athens. Hyper-specific plays like this are discouraged in many other cultural economies,
-including our own, by a desire that scripts be reusable and accessible. A Broadway show
-which spoke deeply to its opening-night audience, but within a year started to alienate
-all newcomers would be considered a failure, but the Athenian model allowed plays to do
-just that.
+\begin{dialogue}
+  \speak{Phoebe} Hey, y'all. I'm Phoebe, and I'd like to welcome you to this production of
+  \theplay{}, \billy{}'s beautiful tragic romance. There will be one 15-minute
+  intermission between the second and third acts. Please silence your phones if you
+  haven't already, and unwrap your snacks now so they don't crinkle during the show.
 
-Being performed as a public service, without a direct financial incentive like ticket
-sales, seems to have encouraged Athenian playwrights to focus more on exposing their
-audiences to new ideas or perspectives, and less on leaving them entertained or happy,
-than the Broadway scene. I think once again of \textit{Ajax}, this time the ambiguous
-light in which it portrays Ajax, one of Athens' founding heroes; Odysseus, arguably the
-most beloved mythical hero; and Athena, the patron goddess of Athens. I'd love to tell you
-what the Greek public thought of \textit{Ajax}, but all I can tell you is that there is no
-record of its having won the City Dionysia. We do know that some very challenging and
-provocative plays won first place, including Aeschylus' \textit{Persians}, which invited
-sympathy for a hated enemy against whom Athens had quite recently fought and won a brutal
-war. This suggests to me that the festival's convoluted judging scheme managed to select
-for some definition of didacticism and artistic integrity, rather than pure entertainment
-value.
+  Before we start, I know the language of \billy{} can be hard to get used to, and people
+  sometimes have a hard time following, so I'll give you a quick plot summary so you know
+  what's going on.
 
-\section*{Imperial Rome}
+  We start with some exposition, to show you that the houses of Montague and Capulet are
+  enemies. It's boring; I'd skip that part if I could. Whatever.
 
-Roman theater, as I understand it, is best summarized by the phrase \enquote{bread and
-  circuses.} Roman plays were funded either by the empire itself or by the patricians, and
-served (with a range of conscious forethought) to placate and distract the plebeians and
-to maintain the social hierarchy. At least as the post-Roman Christians tell it, this
-meant the Romans' theater was gaudy, excessive and meaningless. Their main contribution to
-drama as an art was in importing, preserving and re-performing the works of other
-cultures; every surviving Greek play comes to us through the Romans. If I wanted to be
-particularly political, I might set aside some time to draw parallels with modern America,
-with the message that empire does not promote compelling art. That essay sounds like a
-bummer, though, so instead I'll move on.
+  Romeo of Montague is a thirsty teen, depressed because his crush, Rosaline, doesn't like
+  him back. His pals Mercutio and Benvolio sneak him into a party to cheer him up, where
+  he meets 13-year-old Juliet of Capulet. They fall in love at first sight. \direct{rolls
+    eyes and mimes making themself barf.}
 
-\section*{Medieval religious drama}
+  That night, Romeo sneaks around outside Juliet's house and spies on her a little. He
+  hears her talk about how hot he is, so he does his best to act innocent and pretend he
+  was just passing by, and reveals himself. They get married the next morning.
 
-Frankly, I'm not that interested in this era, which is just as well because there's very
-little to say about it. Medieval European history is characterized by the Church's
-overwhelming influence over art, culture and day-to-day life. The Romans' excesses led the
-Church to reject theater for the better part of a millennium, and when the art does
-re-emerge, we see morality plays like \textit{Everyman}. It seems we actually know very
-little about \textit{Everyman}; we don't know who wrote it, who first performed it, where
-it was performed, or who paid for any of these things. I do feel safe to assert, though,
-that the church was directly involved, either in a positive role by sponsoring the play's
-creation or a negative one in restricting its content. Comparing this period with Broadway
-theater is difficult, as our society is so secular compared to medieval Europe.
+  Then, drama! Fiery Tybalt of Capulet kills Mercutio, Romeo kills Tybalt in revenge, and
+  the prince comes along and exiles Romeo from the city.
 
-\section*{Shakespeare and the renaissance}
+  Juliet decides she doesn't really care that Romeo murdered her cousin, and arranges for
+  him to visit that night for sex, before he leaves town forever.
 
-It's in the renaissance that we first see a theater economy similar to the one I live with
-today, and by Shakespeare's time in the Elizabethan age, the parallels are
-remarkable. Playwrights were commissioned by wealthy patrons, especially the monarchs, and
-used that funding to support initial performances. They earned profit from ticket sales,
-and used the prestige of popular shows to attract patrons for future works. The notable
-difference compared to Broadway is that Shakespeare had no obligation to pay back his
-patrons' gifts; they were not investors. But the model is fundamentally the same:
-monarchs, aristocrats and other people of wealth used financial involvement to dictate the
-content of new plays, and ticket sales provided an incentive to produce popular,
-crowd-pleasing shows.
+  Then, more drama! Juliet's dad arranges a high-society marriage for her, not knowing
+  that Juliet is already married and deflowered. Juliet arranges some convoluted escape
+  plan which involves faking her death. Romeo thinks she's actually dead, and immediately
+  kills himself out of grief. Juliet wakes up, discovers Romeo's corpse, and kills herself
+  for real. Then there's some feel-good stuff where houses Montague and Capulet realize
+  the cost of their feud and make peace, which if you ask me brings us dangerously close
+  to saying \direct{with airquotes} \enquote{hey kid, kill yourself to solve your parents'
+    problems!} But who asked me, right?
 
-Patrons in this era used drama as a means of political conflict and propaganda. On the
-world stage, this takes the form of the monarchs commissioning plays which glorify,
-historicize and mythologize England's past. We can't actually say for certain which of
-Shakespeare's plays were commissioned by Elizabeth or James, or at least, I can't find a
-source which justifies the claim, but we do know that a disproportionate number of
-Shakespeare's plays are devoted to England's frankly very boring history, and that the
-monarchs were his primary patrons. More interesting, on a local level, we see at least one
-instance of an internal dispute using a Shakespeare play to justify an attempted coup: the
-Earl of Essex famously hired Shakespeare's company, the Chamberlain's Men, to perform
-\textit{Richard II} in the hopes of swaying public opinion towards the would-be usurper
-against Elizabeth. (It didn't work.)
+  \direct{checks watch.} Oh, shit, I've spent way too long already. Without further ado,
+  \direct{dons hat and quickly reads the actual prologue}.
+\end{dialogue}
 
-The popular incentive of ticket sales mostly shows through in Shakespeare's dirty jokes
-and reliance on convention. You've already read my take on the dirty jokes, though I'll
-point out that they show up not only in \textit{Romeo and Juliet} but also in otherwise
-serious tragedies; \textit{Theater Histories} cites the porter scene in
-\textit{Macbeth}. In addition to saucy entertainment, audiences expected certain
-conventions, especially in terms of plot and style. Comedies had to end with a marriage,
-and tragedies with a death; \textit{Theater Histories} here points out that \textit{As You
-  Like It} and \textit{Hamlet} had four each of the appropriate ending. Jeremy Lopez has a
-whole book about the conventions expected by theater audiences at this time, and the way
-playwrights (especially Shakespeare) sought to meet those expectations; he highlights puns
-and wordplay, asides and soliloquy, and rapid tonal shifts to keep things exciting. Other
-than the specific plot conventions, which have changed to accommodate a new zeitgeist, I
-think this whole paragraph could apply to Broadway musicals (dramas tend to be more
-experimental and less constrained); they're filled with crowd-pleasing plot moments like
-the long-awaited kiss at the end, dirty jokes, songs which illustrate a character's
-internal monologue, and pretty much every other device Lopez describes.
+I've also considered breaking this summary up into smaller chunks, which I'd read before
+each scene or each act. This could have the added benefits of providing some entertainment
+during set changes, and of further distancing the audience by interrupting their
+immersion.
+
+\section*{Modifications to the script}
+
+If this staging is to convincingly attack the idea that \billy{} wrote \theplay{} as a
+sincere romantic tragedy, it must present his script, unaltered, in its entirety. That
+said, many of \billy{}'s best jokes rely on idioms which no longer make any sense. As a
+result, a narrator will pause the play at strategic points to explain. (I imagine myself
+serving as this narrator, and so I will write my own name in the script.) For example, in
+Act 1, Scene 1, supplementing lines 18--24, I would add:
+
+\begin{dialogue}
+  \speak{Gregory} The quarrel is between our masters and us their men.
+
+  \speak{Sampson} 'Tis all one, I will show myself a tyrant: when I \\
+  have fought with the men, I will be cruel with the \\
+  maids, and cut off their heads.
+
+  \speak{Gregory} The heads of the maids?
+
+  \speak{Sampson} Ay, the heads of the maids, or their maidenheads; \\
+  take it in what sense thou wilt.
+
+  \speak{Phoebe} \direct{while entering from offstage} Could you pause for a moment, guys?
+
+  \direct{to audience} In \billy{}'s time, a \enquote{maiden} is specifically a virgin,
+  and when Sampson talks about \enquote{cutting off their maidenheads,} he means fucking
+  them.
+
+  \direct{beat}
+
+  \direct{to \refer{Sampson} and \refer{Gregory}} Thanks, guys, go ahead. \direct{exits}
+\end{dialogue}
+
+These interruptions would be kept to a minimum, in order to preserve the flow of the
+dialogue. In most cases, strategic emphasis and body language should be sufficient to get
+the point across. The lines directly following make for a good example:
+
+\begin{dialogue}
+  \speak{Gregory} They \direct{chuckles, then continues while wiggling his eyebrows at the
+    audience} must take it in the sense that feel it.
+
+  \speak{Sampson} \direct{mimes holding an exaggerated erect penis} \\
+  Me they shall feel while I am able to stand: and \\
+  'tis known I am a pretty piece of flesh.
+
+  \speak{Gregory} 'Tis well thou art not if; if thou hadst, thou \\
+  hadst been \direct{pauses to produce an ugly wooden fish and hold it to his crotch} poor-John.
+\end{dialogue}
+
+\section*{Setting}
+
+One of the most powerful ways to impose distance between the stage and the audience is to
+add inconsistencies, so that the play lacks a coherent internal world. This is convenient
+because it is easy from a production standpoint, and often reads as comedic.
+
+\theproduction{} will make no effort to ground itself in any one place or time. Visual
+motifs in the set and costumes will be chosen with an eye for what best portrays the
+correct tone or character. Examples to follow.
+
+\section*{Set}
+
+The set will be dirt-cheap, and will look ridiculous. As with all my choices, my goal here
+is to prevent the audience from taking things too seriously.
+
+The only set piece which I see as necessary is the obvious Juliet-style balcony. I haven't
+fully decided just how vulgar I want \theproduction{} to be, but I can certainly imagine a
+version where the Juliet balcony is shaped and painted to look like a vagina, to emphasize
+the sexual subtext in Act 2 Scene 2 and Act 3 Scene 5. A set of pink drapes or blinds
+could make this especially funny, with the drapes closed in Act 2 Scene 2 (before the
+couple has consummated their marriage), and opened in Act 3 Scene 5 (after the fact). A
+wide range of subtler presentations are available, depending on how vulgar I wanted to be;
+anything from a simple white wall with pink finish and pink drapes, all the way to an
+O'Keeffe painting. I suspect the actual decision would be made by what my investors and
+theater administrators would let me get away with.
+
+As to other sets and settings, I imagine:
+
+\begin{itemize}
+\item \enquote{A hall in Capulet's house,} where Romeo and friends crash a party and meet
+  Juliet, will be styled like a 2000s or 2010s high school dance. There will be a disco
+  ball, a punch bowl, and extras in ill-fitting rented suits dancing the \textit{Cha-Cha
+    Slide}. The intention is to say: this is where teenagers go for romance and
+  excitement.
+\item \enquote{Friar Laurence's Cell} will be my best approximation of an authentic monk's
+  cell, with a cot, a bench, a candelabra, an ornate cross and little else. The intention
+  is to say: here is a man whose intense piety has no place alongside the gaudy wealth and
+  teenaged angst of our protagonists.
+\item \enquote{A churchyard; in it a tomb belonging to the Capulets} will be decorated
+  like a cheap haunted house, with plastic gravestones and fake skeletons purchased at a
+  \textit{Spirit Halloween} (or equivalent pop-up store). The intention is to say: this is
+  not a serious encounter with morality, it is a morbid punchline.
+\end{itemize}
+
+\section*{Costumes}
+
+In no particular order:
+
+\begin{itemize}
+\item Prince Escalus and Count Paris, who should appear regal, wealthy and pretentious,
+  will wear medieval finery, with frilly shirts, capes, furs and the like. The prince will
+  wear a crown and carry a scepter. The count will wear a smaller coronet.
+\item Mercutio, who should appear friendly, brash and carefree, will wear jeans and a high
+  school letterman jacket. He will carry an open can of beer at all times.
+\item Lord Capulet, who should appear a stern businessman who knows that violence is bad
+  for profits, will mimic Marlon Brando as the Godfather.
+\item Lady Capulet, similarly, will mimic Morgana King as Mama Corleone.
+\item Lord Montague will mimic Richard Conte as Emilio Barzini, since I think the
+  \textit{Godfather} references are fun.
+\item Fiery Tybalt, who's always looking for a fight, will be a greaser, with a leather
+  jacket, a cigarette and a shiv.
+\item Benvolio, who should appear kind of a wimp, will be a stereotypical nerd. I imagine
+  Crispin Glover as George McFly in \textit{Back to the Future}, with thick glasses,
+  slicked-back hair, and a dress shirt with the top button buttoned but no tie.
+\item Friar Laurence will be a bald monk wearing a brown robe, for similar reasons as my
+  design for his cell.
+\item Romeo, who should appear angsty, young and stupid, will be a goth teen. Dyed
+  jet-black hair, huge steel ear studs, a thin goatee, tight black ripped jeans and a tee
+  shirt.
+\item Juliet, who should appear young, stupid and naïve, will have long wavy brown hair, a
+  pretty blue dress, and lots of blush.
+\end{itemize}

### 2022-03-18 paper.tex
midterm paper for theater and medicine

@@
-\newcommand{\theplay}{\textit{Romeo and Juliet}}
-\newcommand{\billy}{Shakespeare}
-\newcommand{\theproduction}{\textit{Slapstick Shakespeare}}
+\newcommand{\lear}{\textit{King Lear}}
+\newcommand{\oedipus}{\textit{Oedipus Tyrannus}}
+% \section*{Tell 'em what you're gonna tell 'em}
+Shakespeare's \lear{} and Sophocles' \oedipus{}, despite being separated by two
+millenia and situated in vastly different political systems, show similar
+outlooks on the health and fitness of the monarch, of the ruling family, and of
+the state. Both plays are concerned with the feedback loop between the stress of
+rulership, the monarch's mental health, and the disastrous ripples caused by the
+monarch's short-sighted decision-making. Sophocles' more reserved, Aristotelian
+dramatic form embodies these concerns almost exclusively in Oedipus himself,
+while Shakespeare's greater complexity allows him to spread the embodiments of
+illness between Lear and Gloucester. Sophocles' style is more ambiguous,
+inviting the audience to empathize with Oedipus even as we watch him descend
+into madness and take Thebes with him, whereas Shakespeare is content to have us
+dislike Lear, with the tragedy of the play coming not from his personal
+suffering but the suffering he inflicts on those around him. Still, the two
+plays show remarkable consistency in their depictions of impulsivity,
+self-assuredness and rage and their linking of those traits to the suffering and
+downfall of the state.
 
-I read \theplay{} in high school, led by an English teacher who frankly wasn't very
-good. She taught the play as a serious tragic romance. I experienced it more as a dark
-comedy; an elaborate farce at the expense of young people. I was aided by a good annotated
-edition (I've unfortunately been unable to find the same one since), which explained all
-the dirty jokes.
+% \section*{Tell 'em}
 
-The plot of \theplay{} always struck me as being... less than romantic. The supposed
-\enquote{star-crossed lovers} are a pair of very young teenagers (Juliet is 13) who become
-instantly infatuated upon meeting at a party. In fact, it's not even the first time Romeo
-has fallen hopelessly in love! I'd like to stage the play in such a way to highlight the
-absurdity, bring the raunchy jokes to a forefront, and hopefully destroy the canonical
-reading that it's a sincere romance.
+% \subsection*{Impulsivity, self-assurance, rage}
 
-The dramatic literature major in me feels obligated to describe my approach as a comedic
-\textit{Verfremdungseffekt}. Similar to how Brechtian drama imposes distance in order to
-force its audience to critically reconsider the social and political circumstances
-presented, \theproduction{} will force its audience to critically reconsider the canonical
-interpretation of \theplay{}, particularly in terms of genre and tone.
+At the beginning of \oedipus{}, the titular tyrant is quite lucid; it is only as
+the play progresses and his fate becomes clear that his mental state is called
+into question. He first suffers a burst of rage when Tireseas refuses to reveal
+the identity of Laius' killer; Oedipus' inflamed response is,
+\textquote[{\cite[ll. 334--335, 339--340]{oedipus}}]{Nothing? Damn you! You
+  could make the coldest stone // Burn with rage. \textelp{} Could anyone not be
+  angry after hearing // How you hold our city in such contempt?} Though
+Tireseas is confoundingly unhelpful, he is not wrong to refuse Oedipus' demands,
+nor is Oedipus wrong to be enraged by them. Sophocles shows us that Oedipus, in
+the course of his duties as tyrant, is unavoidably exposed to circumstances
+which are stressful beyond human tolerance.
 
-\section*{A new prologue}
+Lear, on the other hand, has already succumbed to that stress by the time his
+play begins. Instead of watching his descent from lucidity to madness, we are
+forced to reconstruct an image of the king he once was by contrasting his older
+advisors' loyalty to him against the younger generation's eagerness to oust
+him. Kent, after being banished and threatened with death for opposing Lear
+\parencite[1.1]{lear}, returns in disguise to continue his service to the
+king. Lear is justifiably confused that an apparent stranger would offer himself
+into the service of a destitute monarch, and the audience is similarly confused
+that Kent would return to his thankless and dangerous work. In an aside, Kent
+explains to us, \textquote[{\cite[1.4]{lear}}]{thy master, whom thou lovest, //
+  Shall find thee full of labours}. To Lear, Kent gives us a glimpse of the
+traits which inspired that love, swearing \textquote[{\cite[1.4]{lear}}]{to
+  serve him truly that will put me in trust: to love him that is honest; to
+  converse with him that is wise \textelp{}}. A somewhat contrived turn of
+dialogue has Lear ask Kent his age, and we are told that Kent is 48. This scene
+shows us Lear's former glory, the years Kent spent in service of an honest,
+wise, just king. We cannot help but be struck by the contrast between Kent's
+description and the Lear who stands before us onstage.
 
-Rather than a chorus, I think I'll read the two prologues (there's one before each of the
-first two acts), wearing a mobcap with a feather, to signify my being in-character as the
-Bard. Before the first prologue, I'll enter holding the hat, to signify being
-out-of-character, and address the audience with an introduction to set the tone:
+Goneril draws our attention to the transformation again later in the same scene,
+entreating her father, \textquote[{\cite[1.4]{lear}}]{I would you would make use
+  of that good wisdom \textelp{} and put away // These dispositions, that of
+  late transform you // From what you rightly are}. Lear, of course, objects,
+insisting that he is now as he's always been, but we know that is not the
+case. The stress of rule has taken its toll on Lear, and he is no longer the man
+he once was.
 
+The symptoms displayed by Lear and Oedipus bear highlighting: both are prone to
+paranoia and fits of rage. Oedipus sees schemes against himself, Thebes and
+Laius around every corner; he accuses Tireseas of having part in Laius' death
+\parencite[ll. 345--349]{oedipus}, Creon of conspiring to falsely accuse Oedipus
+and claim power for himself \parencite[ll. 385--389, 399--402,
+572--573]{oedipus}, and espouses a philosophy of rule based in mutual
+conspiracy, \textquote[{\cite[ll. 618--621]{oedipus}}]{When a conspirator moves
+  quickly against me, // Then I must be quick to conspire back at him. // If I
+  hesitate I will lose the initiative, // While he will seize the moment and
+  strike}. Sophocles, relentlessly didactic, does not place the onus on us to
+interpret and reject Oedipus' paranoia, but has the chorus tell us,
+\textquote[{\cite[ll. 616--617]{oedipus}}]{Anyone who treads carefully will
+  agree: // Decisions made too quickly are dangerous}. Oedipus then demands that
+Creon (who is portrayed in \textit{oedipus} as noble, just and content, with no
+lust for power) be put to death, leading to a fast-paced dialogue revelatory of
+Oedipus' state:
+
+\begin{minipage}[t]{0.8\linewidth}
   \begin{dialogue}
-  \speak{Phoebe} Hey, y'all. I'm Phoebe, and I'd like to welcome you to this production of
-  \theplay{}, \billy{}'s beautiful tragic romance. There will be one 15-minute
-  intermission between the second and third acts. Please silence your phones if you
-  haven't already, and unwrap your snacks now so they don't crinkle during the show.
+    \speak{Creon} And you could be wrong.
+    \speak{Oedipus} Nevertheless, I must rule.
+    \speak{Creon} Not if it means ruling badly.
+    \speak{Oedipus} Oh, my city, my city\ldots{}
+    \speak{Creon} I belong in this city, it's not only yours!
 
-  Before we start, I know the language of \billy{} can be hard to get used to, and people
-  sometimes have a hard time following, so I'll give you a quick plot summary so you know
-  what's going on.
+    \attrib{\cite[ll. 128--130]{oedipus}}
+  \end{dialogue}
+\end{minipage}
 
-  We start with some exposition, to show you that the houses of Montague and Capulet are
-  enemies. It's boring; I'd skip that part if I could. Whatever.
+\vspace{0.5\baselineskip}
 
-  Romeo of Montague is a thirsty teen, depressed because his crush, Rosaline, doesn't like
-  him back. His pals Mercutio and Benvolio sneak him into a party to cheer him up, where
-  he meets 13-year-old Juliet of Capulet. They fall in love at first sight. \direct{rolls
-    eyes and mimes making themself barf.}
+In this moment, we see that the stresses of ruling (and espeically, of trying to
+fix the plague caused by his rule) have driven Oedipus from a just tyranny to
+the violence and oppression described by Plato. Oedipus feels pressured to make
+quick, firm decisions even when they are wrong, and sees the city as his
+possession rather than the object of his service. What is worse, we do not hate
+Oedipus for this denigration any more than Kent hates Lear; it is the natural
+consequence of circumstances outside Oedipus' control.
 
-  That night, Romeo sneaks around outside Juliet's house and spies on her a little. He
-  hears her talk about how hot he is, so he does his best to act innocent and pretend he
-  was just passing by, and reveals himself. They get married the next morning.
+% \subsection*{Suffering and downfall of the state}
 
-  Then, drama! Fiery Tybalt of Capulet kills Mercutio, Romeo kills Tybalt in revenge, and
-  the prince comes along and exiles Romeo from the city.
+Both \lear{} and \oedipus{} are keen to show us that, because of their
+responsibilities to their respective states, the tragic suffering of the titular
+rulers extends to their subjects. For Oedipus, this manifests as a plague, while
+for Lear it is a war. (I am tempted to investigate these different portrayals of
+danger to the state in search ofsome statement about the political circumstances
+of Athens in 429 BCE and England in 1606, but that seems at best tangentially
+related to the argument I'm making here.) Again, the structures of these plays
+lead us to experience the suffering differently; Sophocles' reliance on a
+well-known myth, his limited scope and his character-driven focus places us
+already in the plague at the start of the play, without depicting its
+beginnings. We are shown, however, the gravity and horror of the plague
+\parencite[e.g. ll. 22--30]{oedipus}, and we are reminded that the plague is a
+divine punishment for Oedipus' hamartia, his killing of Laius
+\parencite[ll. 101--110]{oedipus}. Much later, Sophocles eplicicitly links that
+killing to Oedipus' temper \parencite[ll. 807--813]{oedipus}. After, quite
+frankly, a very long journey, our understanding of events is: Oedipus, in a
+(provoked and arguably justified) fit of rage, killed Laius, then, without
+realizing, claimed his throne and house. The gods, outraged that Thebes would,
+also without realizing, accept the killer of their rightful king and crown him
+tyrant, strike the city with a plague. All of which is to say that the Theban
+plague was caused by Oedipus' loss of control over his temper.
 
-  Juliet decides she doesn't really care that Romeo murdered her cousin, and arranges for
-  him to visit that night for sex, before he leaves town forever.
+% lear causes a war, lmao
 
-  Then, more drama! Juliet's dad arranges a high-society marriage for her, not knowing
-  that Juliet is already married and deflowered. Juliet arranges some convoluted escape
-  plan which involves faking her death. Romeo thinks she's actually dead, and immediately
-  kills himself out of grief. Juliet wakes up, discovers Romeo's corpse, and kills herself
-  for real. Then there's some feel-good stuff where houses Montague and Capulet realize
-  the cost of their feud and make peace, which if you ask me brings us dangerously close
-  to saying \direct{with airquotes} \enquote{hey kid, kill yourself to solve your parents'
-    problems!} But who asked me, right?
+At this point in the paper, I had intended to use textal evidence from \lear{}
+to convince you that Lear's mental deterioration caused the war between the
+English and the French, but honestly, it seems unnecessary. That reading is so
+obviously on the surface of this play that I won't waste either of our time
+beating a dead horse.
 
-  \direct{checks watch.} Oh, shit, I've spent way too long already. Without further ado,
-  \direct{dons hat and quickly reads the actual prologue}.
-\end{dialogue}
+% \subsection*{Blindness}
 
-I've also considered breaking this summary up into smaller chunks, which I'd read before
-each scene or each act. This could have the added benefits of providing some entertainment
-during set changes, and of further distancing the audience by interrupting their
-immersion.
+% \section*{Tell 'em what you told 'em}
 
-\section*{Modifications to the script}
-
-If this staging is to convincingly attack the idea that \billy{} wrote \theplay{} as a
-sincere romantic tragedy, it must present his script, unaltered, in its entirety. That
-said, many of \billy{}'s best jokes rely on idioms which no longer make any sense. As a
-result, a narrator will pause the play at strategic points to explain. (I imagine myself
-serving as this narrator, and so I will write my own name in the script.) For example, in
-Act 1, Scene 1, supplementing lines 18--24, I would add:
-
-\begin{dialogue}
-  \speak{Gregory} The quarrel is between our masters and us their men.
-
-  \speak{Sampson} 'Tis all one, I will show myself a tyrant: when I \\
-  have fought with the men, I will be cruel with the \\
-  maids, and cut off their heads.
-
-  \speak{Gregory} The heads of the maids?
-
-  \speak{Sampson} Ay, the heads of the maids, or their maidenheads; \\
-  take it in what sense thou wilt.
-
-  \speak{Phoebe} \direct{while entering from offstage} Could you pause for a moment, guys?
-
-  \direct{to audience} In \billy{}'s time, a \enquote{maiden} is specifically a virgin,
-  and when Sampson talks about \enquote{cutting off their maidenheads,} he means fucking
-  them.
-
-  \direct{beat}
-
-  \direct{to \refer{Sampson} and \refer{Gregory}} Thanks, guys, go ahead. \direct{exits}
-\end{dialogue}
-
-These interruptions would be kept to a minimum, in order to preserve the flow of the
-dialogue. In most cases, strategic emphasis and body language should be sufficient to get
-the point across. The lines directly following make for a good example:
-
-\begin{dialogue}
-  \speak{Gregory} They \direct{chuckles, then continues while wiggling his eyebrows at the
-    audience} must take it in the sense that feel it.
-
-  \speak{Sampson} \direct{mimes holding an exaggerated erect penis} \\
-  Me they shall feel while I am able to stand: and \\
-  'tis known I am a pretty piece of flesh.
-
-  \speak{Gregory} 'Tis well thou art not if; if thou hadst, thou \\
-  hadst been \direct{pauses to produce an ugly wooden fish and hold it to his crotch} poor-John.
-\end{dialogue}
-
-\section*{Setting}
-
-One of the most powerful ways to impose distance between the stage and the audience is to
-add inconsistencies, so that the play lacks a coherent internal world. This is convenient
-because it is easy from a production standpoint, and often reads as comedic.
-
-\theproduction{} will make no effort to ground itself in any one place or time. Visual
-motifs in the set and costumes will be chosen with an eye for what best portrays the
-correct tone or character. Examples to follow.
-
-\section*{Set}
-
-The set will be dirt-cheap, and will look ridiculous. As with all my choices, my goal here
-is to prevent the audience from taking things too seriously.
-
-The only set piece which I see as necessary is the obvious Juliet-style balcony. I haven't
-fully decided just how vulgar I want \theproduction{} to be, but I can certainly imagine a
-version where the Juliet balcony is shaped and painted to look like a vagina, to emphasize
-the sexual subtext in Act 2 Scene 2 and Act 3 Scene 5. A set of pink drapes or blinds
-could make this especially funny, with the drapes closed in Act 2 Scene 2 (before the
-couple has consummated their marriage), and opened in Act 3 Scene 5 (after the fact). A
-wide range of subtler presentations are available, depending on how vulgar I wanted to be;
-anything from a simple white wall with pink finish and pink drapes, all the way to an
-O'Keeffe painting. I suspect the actual decision would be made by what my investors and
-theater administrators would let me get away with.
-
-As to other sets and settings, I imagine:
-
-\begin{itemize}
-\item \enquote{A hall in Capulet's house,} where Romeo and friends crash a party and meet
-  Juliet, will be styled like a 2000s or 2010s high school dance. There will be a disco
-  ball, a punch bowl, and extras in ill-fitting rented suits dancing the \textit{Cha-Cha
-    Slide}. The intention is to say: this is where teenagers go for romance and
-  excitement.
-\item \enquote{Friar Laurence's Cell} will be my best approximation of an authentic monk's
-  cell, with a cot, a bench, a candelabra, an ornate cross and little else. The intention
-  is to say: here is a man whose intense piety has no place alongside the gaudy wealth and
-  teenaged angst of our protagonists.
-\item \enquote{A churchyard; in it a tomb belonging to the Capulets} will be decorated
-  like a cheap haunted house, with plastic gravestones and fake skeletons purchased at a
-  \textit{Spirit Halloween} (or equivalent pop-up store). The intention is to say: this is
-  not a serious encounter with morality, it is a morbid punchline.
-\end{itemize}
-
-\section*{Costumes}
-
-In no particular order:
-
-\begin{itemize}
-\item Prince Escalus and Count Paris, who should appear regal, wealthy and pretentious,
-  will wear medieval finery, with frilly shirts, capes, furs and the like. The prince will
-  wear a crown and carry a scepter. The count will wear a smaller coronet.
-\item Mercutio, who should appear friendly, brash and carefree, will wear jeans and a high
-  school letterman jacket. He will carry an open can of beer at all times.
-\item Lord Capulet, who should appear a stern businessman who knows that violence is bad
-  for profits, will mimic Marlon Brando as the Godfather.
-\item Lady Capulet, similarly, will mimic Morgana King as Mama Corleone.
-\item Lord Montague will mimic Richard Conte as Emilio Barzini, since I think the
-  \textit{Godfather} references are fun.
-\item Fiery Tybalt, who's always looking for a fight, will be a greaser, with a leather
-  jacket, a cigarette and a shiv.
-\item Benvolio, who should appear kind of a wimp, will be a stereotypical nerd. I imagine
-  Crispin Glover as George McFly in \textit{Back to the Future}, with thick glasses,
-  slicked-back hair, and a dress shirt with the top button buttoned but no tie.
-\item Friar Laurence will be a bald monk wearing a brown robe, for similar reasons as my
-  design for his cell.
-\item Romeo, who should appear angsty, young and stupid, will be a goth teen. Dyed
-  jet-black hair, huge steel ear studs, a thin goatee, tight black ripped jeans and a tee
-  shirt.
-\item Juliet, who should appear young, stupid and naïve, will have long wavy brown hair, a
-  pretty blue dress, and lots of blush.
-\end{itemize}
+Shakespeare's \lear{} and Sophocles' \oedipus{} both present the dangers
+inherint in absolute rule: the ruler is placed under incredible stress, and a
+ruler who succumbs to stress will cause great harm not just to themself but to
+the state and all its citizens. There are, in my mind, two potential responses
+to this problem, though neither author explicitly suggests either. In a
+monarchist reading, the consequence is that kings (and tyrants) must be
+exceptionally steady, level-headed people, and must be surrounded by strong,
+trusted advisors who are empowered to stop the ruler from making poor
+choices. Both plays show a hurdles in this solution: Lear and Oedipus are both
+professed to be the steady, level-headed type, and both are surrounded by
+trusted advisors; neither of these safeguards prevents the tragedy of the plays
+from unfolding. The antimonarchist reading is, of course, that divesting
+absolute rule in a single person is always a losing proposition, no matter how
+well-suited that ruler appears, and so political systems which distribute power
+among multiple people must be preferred.

### 2022-03-23 paper.tex
an essay that should've been longer but wasn't

@@
-\newcommand{\lear}{\textit{King Lear}}
-\newcommand{\oedipus}{\textit{Oedipus Tyrannus}}
-% \section*{Tell 'em what you're gonna tell 'em}
-Shakespeare's \lear{} and Sophocles' \oedipus{}, despite being separated by two
-millenia and situated in vastly different political systems, show similar
-outlooks on the health and fitness of the monarch, of the ruling family, and of
-the state. Both plays are concerned with the feedback loop between the stress of
-rulership, the monarch's mental health, and the disastrous ripples caused by the
-monarch's short-sighted decision-making. Sophocles' more reserved, Aristotelian
-dramatic form embodies these concerns almost exclusively in Oedipus himself,
-while Shakespeare's greater complexity allows him to spread the embodiments of
-illness between Lear and Gloucester. Sophocles' style is more ambiguous,
-inviting the audience to empathize with Oedipus even as we watch him descend
-into madness and take Thebes with him, whereas Shakespeare is content to have us
-dislike Lear, with the tragedy of the play coming not from his personal
-suffering but the suffering he inflicts on those around him. Still, the two
-plays show remarkable consistency in their depictions of impulsivity,
-self-assuredness and rage and their linking of those traits to the suffering and
-downfall of the state.
+\newcommand{\huter}{\textit{Humanism and Terror}}
+\newcommand{\dark}{\textit{Darkness at Noon}}
 
-% \section*{Tell 'em}
+In his essay \huter{}, Maurice Merleau-Ponty presents an ultimately unconvincing
+counterpoint to Arthur Koestler's fictionalization of the Soviet show-trials,
+\dark{}. Ponty insists that his essay is not intended as a defense of the Soviet regime's
+violent practices under Stalin, but rather to inspect and conceptualize them in Marxist
+terms. In Ponty's view, Koestler's novel presents a fundamentally liberal critique of
+communist violence, and therefore fails to relate itself to the realities of the Soviet
+situation. Ponty's communist frameworks for understanding the novel, however, contain
+flaws and inconsistencies which cannot be internally reconciled. As such, external
+critique like Koestler's are not only reasonable but necessary. In fact, Ponty's attempt
+to reconcile the internal minds of Koestler's characters and their actions with Marxist
+philosophy completely misses the point of \dark{}, that interpreting and effecting Marxist
+ideology led the Soviets to beliefs and actions irreconcilable with that ideology.
 
-% \subsection*{Impulsivity, self-assurance, rage}
+% pick apart various complaints of ponty's
 
-At the beginning of \oedipus{}, the titular tyrant is quite lucid; it is only as
-the play progresses and his fate becomes clear that his mental state is called
-into question. He first suffers a burst of rage when Tireseas refuses to reveal
-the identity of Laius' killer; Oedipus' inflamed response is,
-\textquote[{\cite[ll. 334--335, 339--340]{oedipus}}]{Nothing? Damn you! You
-  could make the coldest stone // Burn with rage. \textelp{} Could anyone not be
-  angry after hearing // How you hold our city in such contempt?} Though
-Tireseas is confoundingly unhelpful, he is not wrong to refuse Oedipus' demands,
-nor is Oedipus wrong to be enraged by them. Sophocles shows us that Oedipus, in
-the course of his duties as tyrant, is unavoidably exposed to circumstances
-which are stressful beyond human tolerance.
+%% rubashov's beliefs aren't acutally marxist
+%%% well, duh
+Fundamental to Ponty's critique of \dark{} is his equating Rubashov's understanding of
+Soviet doctrine with Koestler's understanding of Marxism. Referencing a moment from the
+end of the novel, when a defeated and fatalistic Rubashov reflects on the relationship
+between the individual and the collective \parencite[77]{koestler}, Ponty writes,
+\textquote[{\cite[23]{ponty}}]{who said that history is a clockwork and the individual a
+  wheel? It was not Marx; it was Koestler. It is strange that in Koestler there is no
+  inkling of the commonplace notion \textelp{} Koestler has never given much thought to
+  the simple idea of a dialectic in history}. As a self-proclaimed dramaturg, it is
+difficult for me to express my indignation at Ponty's reading without vastly and rudely
+departing from the tonal and stylistic conventions of a college essay. In the context of
+\dark{}'s third-person limited narration, any halfway-literate reader understands that the
+passage quoted by Ponty is a representation of Rubashov's subjective state of mind, not an
+expression of Koestler's beliefs. I have tried to excuse this lapse of Ponty's to myself,
+wondering if perhaps the series of translations and second languages through which he must
+have experienced the novel may have occluded the meaning which is so obvious to me,
+whether German, English and French literary conventions may be so different as to have led
+Ponty astray, but Koestler goes as far as to introduce the passage,
+\textquote[{\cite[77]{koestler}}]{Rubashov wandered through his cell. In old days he would
+  have shamefacedly denied himself this sort of childish musing. Now he was not
+  ashamed}. Ponty's critique here depends on conflating Koestler's own beliefs with the
+\enquote{childish musings} he ascribes to a fictional character. Any high school
+literature student can see that these two are not interchangeable. I refuse to believe
+that Ponty lacks such basic reading comprehension, and so I must conclude that he is
+knowingly projecting onto Koestler a misunderstanding which he can attack, a
+misunderstanding that Koestler has not shown any signs of suffering.
 
-Lear, on the other hand, has already succumbed to that stress by the time his
-play begins. Instead of watching his descent from lucidity to madness, we are
-forced to reconstruct an image of the king he once was by contrasting his older
-advisors' loyalty to him against the younger generation's eagerness to oust
-him. Kent, after being banished and threatened with death for opposing Lear
-\parencite[1.1]{lear}, returns in disguise to continue his service to the
-king. Lear is justifiably confused that an apparent stranger would offer himself
-into the service of a destitute monarch, and the audience is similarly confused
-that Kent would return to his thankless and dangerous work. In an aside, Kent
-explains to us, \textquote[{\cite[1.4]{lear}}]{thy master, whom thou lovest, //
-  Shall find thee full of labours}. To Lear, Kent gives us a glimpse of the
-traits which inspired that love, swearing \textquote[{\cite[1.4]{lear}}]{to
-  serve him truly that will put me in trust: to love him that is honest; to
-  converse with him that is wise \textelp{}}. A somewhat contrived turn of
-dialogue has Lear ask Kent his age, and we are told that Kent is 48. This scene
-shows us Lear's former glory, the years Kent spent in service of an honest,
-wise, just king. We cannot help but be struck by the contrast between Kent's
-description and the Lear who stands before us onstage.
+%%% rubashov's view of history is not the same as marx's, but it is (or at least may be)
+%%% that of the soviet communists
 
-Goneril draws our attention to the transformation again later in the same scene,
-entreating her father, \textquote[{\cite[1.4]{lear}}]{I would you would make use
-  of that good wisdom \textelp{} and put away // These dispositions, that of
-  late transform you // From what you rightly are}. Lear, of course, objects,
-insisting that he is now as he's always been, but we know that is not the
-case. The stress of rule has taken its toll on Lear, and he is no longer the man
-he once was.
+We see this same flaw in Ponty's criticism of Koestler's supposed outlook on
+history. Ponty reproduces several passages in which Rubashov characterizes History (with a
+capital \enquote{H}) as a matter of absolute truth, a force beyond comprehension, and a
+linear march from the darkness of the past to a bright future \parencite[19]{ponty}. There
+are many more moments in \dark{} than just the ones Ponty cites; of these, perhaps most
+striking is when Rubashov reflects on excommunicating Richard, having told him,
+\textquote[{\cite[14--15]{koestler}}]{The Party can never be mistaken. \textelp{} The
+Party is the embodiment of the revolutionary idea in history. History knows no scruples
+and no hesitation. Inert and unerring, she flows towards her goal. \textelp{} History
+knows her way. She makes no mistakes. He who has not absolute faith in History does not
+belong in the Party's ranks}. Ponty attacks this conception of history both on the grounds
+that it is contrary to Marx's conception of \textit{praxis} \parencite[18]{ponty}, as well
+as in its own right as a fragile idea which insulates people from their agency
+\parencite[19]{ponty}. In these critiques, Ponty is exactly correct. But, despite managing
+at least to attribute these thoughts to Rubashov rather than Koestler, Ponty seems to
+believe (in fact, Ponty makes this belief quite explicit) that Koestler perceives history
+this way, or that Koestler is claiming that this is a Marxist perception of
+history \parencite[20]{ponty}. Neither is the case, or at least, neither is evidenced by
+\dark{}. Rather, Koestler's claim is that the Soviets held this conception of history,
+that the Soviets misinterpreted Marx this way. Ponty gets so, so close to critically
+engaging with this claim, writing, \textquote[{\cite[23--24]{ponty}}]{\textelp{} if the
+alternatives of subjectivism and objectivism are resolved in Marx's Marxism, the question
+still remains whether this is so in communism as a reality and whether the majority of
+Communists believe in incorporating subjectivity}. But Ponty doesn't actually grapple with
+the question posed by \dark{}, namely, whether the leaders the U.S.S.R. understand the
+humanist ideals upon which their state is supposedly founded and whether they translate
+those ideals into moral actions.
 
-The symptoms displayed by Lear and Oedipus bear highlighting: both are prone to
-paranoia and fits of rage. Oedipus sees schemes against himself, Thebes and
-Laius around every corner; he accuses Tireseas of having part in Laius' death
-\parencite[ll. 345--349]{oedipus}, Creon of conspiring to falsely accuse Oedipus
-and claim power for himself \parencite[ll. 385--389, 399--402,
-572--573]{oedipus}, and espouses a philosophy of rule based in mutual
-conspiracy, \textquote[{\cite[ll. 618--621]{oedipus}}]{When a conspirator moves
-  quickly against me, // Then I must be quick to conspire back at him. // If I
-  hesitate I will lose the initiative, // While he will seize the moment and
-  strike}. Sophocles, relentlessly didactic, does not place the onus on us to
-interpret and reject Oedipus' paranoia, but has the chorus tell us,
-\textquote[{\cite[ll. 616--617]{oedipus}}]{Anyone who treads carefully will
-  agree: // Decisions made too quickly are dangerous}. Oedipus then demands that
-Creon (who is portrayed in \textit{oedipus} as noble, just and content, with no
-lust for power) be put to death, leading to a fast-paced dialogue revelatory of
-Oedipus' state:
+In fact, Ponty acknowledges in his preface that the Soviets do not live up to their goals,
+writing, \textquote[{\cite[xx]{ponty}}]{In the last ten years in the U.S.S.R. the social
+hierarchy has become considerably accentuated. The proletariat plays an insignificant role
+in the Party Congresses. \textelp{} There is an increasing difference between what
+Communists think and what they write because there is a widening gap between their
+intentions and their deeds}. In this moment, Ponty gives up the game: he admits, before
+even beginning his essay proper, that Koestler's depiction of Soviet politics is
+accurate. Ponty even demonstrates his ability to understand other works which have the
+same goal as \dark{}, framing a quotation from Raymond Aron's \textit{Essai sur la théorie
+de l'histoire dans l'allemagne contempoiraine la philosophie critique de l'histoire},
+\textquote[{\cite[xli]{ponty}}]{\textins{Aron} was able to express a view that was not his
+own but which he considered to be at least one of the more serious views without anyone
+accusing him of toadying Nazi or Communist power}. And yet, Ponty's essay proper labors
+under the assumption that Koestler was writing to portray his own conceptualization of the
+philosophy of Marx, rather than, as Ponty himself puts it, \enquote{to express a view that
+was not his own.}
 
-\begin{minipage}[t]{0.8\linewidth}
-  \begin{dialogue}
-    \speak{Creon} And you could be wrong.
-    \speak{Oedipus} Nevertheless, I must rule.
-    \speak{Creon} Not if it means ruling badly.
-    \speak{Oedipus} Oh, my city, my city\ldots{}
-    \speak{Creon} I belong in this city, it's not only yours!
+%%% "but koestler is putting all these thoughts in his book, even if they're fictional,
+%%% doesn't that constitute endorsement?"
 
-    \attrib{\cite[ll. 128--130]{oedipus}}
-  \end{dialogue}
-\end{minipage}
+%%% no, actually, rubashov is a dick and an idiot, and we're not particularly supposed to
+%%% like or agree with him, just to understand who he is and why he thinks the way he
+%%% does. the whole point of the book is "how did the soviet union produce people like
+%%% rubashov, and how do people like rubashov understand the world?"
 
-\vspace{0.5\baselineskip}
+%% marxist history, violence in history, violence against violence in history, and
+%% violence in pursuit of peace
 
-In this moment, we see that the stresses of ruling (and espeically, of trying to
-fix the plague caused by his rule) have driven Oedipus from a just tyranny to
-the violence and oppression described by Plato. Oedipus feels pressured to make
-quick, firm decisions even when they are wrong, and sees the city as his
-possession rather than the object of his service. What is worse, we do not hate
-Oedipus for this denigration any more than Kent hates Lear; it is the natural
-consequence of circumstances outside Oedipus' control.
+%%% contradiction between requiring humanist critiques but ignoring the humanism in
+%%% koestler's critique
 
-% \subsection*{Suffering and downfall of the state}
+One of Ponty's more infuriating critiques of \dark{} is that Koestler, apparently, is
+discussing Soviet politics from a liberal perspective. As Ponty puts it, in order to
+\textquote[{\cite[xvii]{ponty}}]{understand the Communist problem, it is necessary to
+start by placing the Moscow Trials in the revolutionary \textit{Stimmung} of violence
+apart from which they are inconceivable}. (Why O'Neill chose to reproduce the German word
+\textit{Stimmung} rather than translate it as \enquote{context} or \enquote{atmosphere} I
+do not know.) Setting aside the irony that Ponty insists we engage with communism on its
+terms while himself refusing to engage with \dark{} on its, Ponty's critique here is
+frustrating because \dark{} represents Koestler's attempt to place the Moscow Trials into
+the violent revolutionary atmosphere of the U.S.S.R., and to inhabit the mind of a
+protagonist who breathed it. It is for this purpose that we see Rubashov reflecting on
+undertaking violent revolutionary acts. When informing Richard that he is to be ejected
+from the Party, especially, we experience Rubashov's violence. In the sequence, we watch
+Rubashov expertly and routinely evade fascist operatives, and we experience his
+frustration that Richard becomes intimidated by the presence of a soldier near their
+meeting place, risking their cover \parencite[13]{koestler}. We are shown that, without
+Richard's knowledge, another communist in his cell has been passing information to
+Rubashov, and to Rubashov this is normal \parencite[13]{koestler}. We see both mens'
+impassivity as Richard reports to Rubashov the number of communists who have been arrested
+and presumably killed, including Richard's wife, and we learn that this is an expected
+hazard of the job \parencite[13]{koestler}. We listen as Rubashov explains how the
+leaflets Richard has been deseminating have undermined the Party, and though we as readers
+find Richard's message less objectionable than the official version, our sense is that
+Rubashov, having lived in the world we are being shown, believes in the necessity of the
+Party's actions and messages, and believes that Richard is a genuine threat to the
+revolution \parencite[14]{koestler}. Richard's reaction makes clear that what Rubashov is
+doing is violent; Richard begs, \textquote[{\cite[16]{koestler}}]{You c-can't throw me to
+the wolves, c-comrade}. Rubashov, who believes that his revolutionary work is more
+important that Richard's life, shows the man no sympathy, and though he does experience
+regrets at the end of the sequence, his final thought on the matter is that
+\textquote[{\cite[16]{koestler}}]{the affair with Richard had to be concluded}. In short,
+Koestler makes a concerted effort in \dark{} to place us in Ponty's
+\enquote{\textit{Stimmung} of violence.}
 
-Both \lear{} and \oedipus{} are keen to show us that, because of their
-responsibilities to their respective states, the tragic suffering of the titular
-rulers extends to their subjects. For Oedipus, this manifests as a plague, while
-for Lear it is a war. (I am tempted to investigate these different portrayals of
-danger to the state in search ofsome statement about the political circumstances
-of Athens in 429 BCE and England in 1606, but that seems at best tangentially
-related to the argument I'm making here.) Again, the structures of these plays
-lead us to experience the suffering differently; Sophocles' reliance on a
-well-known myth, his limited scope and his character-driven focus places us
-already in the plague at the start of the play, without depicting its
-beginnings. We are shown, however, the gravity and horror of the plague
-\parencite[e.g. ll. 22--30]{oedipus}, and we are reminded that the plague is a
-divine punishment for Oedipus' hamartia, his killing of Laius
-\parencite[ll. 101--110]{oedipus}. Much later, Sophocles eplicicitly links that
-killing to Oedipus' temper \parencite[ll. 807--813]{oedipus}. After, quite
-frankly, a very long journey, our understanding of events is: Oedipus, in a
-(provoked and arguably justified) fit of rage, killed Laius, then, without
-realizing, claimed his throne and house. The gods, outraged that Thebes would,
-also without realizing, accept the killer of their rightful king and crown him
-tyrant, strike the city with a plague. All of which is to say that the Theban
-plague was caused by Oedipus' loss of control over his temper.
+%%% highlight is questions about violence between "now" and "lenin's day"; compare to
+%%% lenin's hanging order
 
-% lear causes a war, lmao
+Koestler's Rubashov makes several assumptions about the atmosphere of violence in which he
+lives, and Ponty does not question these assumptions. Key among these is the assumption
+that the Soviets under Lenin were just, and used violence only where necessary or useful
+to advance the Revolution. In this version of history, the needless and excessive violence
+which Koestler and Ponty both decry began after Lenin's death and Stalin's rise to
+power. Ponty makes this assumption explicit, asking,
+\textquote[{\cite[xviii]{ponty}}]{Does the violence in today's communism have the same
+sense it had in Lenin's day?} In Koestler, who avoids using the names of Soviet officials
+(this choice is interesting and worth investigating, but outside the scope of this essay),
+the assumption is not tied directly to Lenin, but still revealed in Rubashov's linking
+memories to years. He reminisces, \textquote[{\cite[11]{koestler}}]{It was in the year
+1933, during the first months of terror \textelp{} The Party was no longer a political
+organization; it was nothing but a thousand-armed and thousand-headed mass of bleeding
+flesh}. This timeline of the Party's descent by 1933 lines up quite nicely with Lenin's
+death in 1924 and Stalin's rise to absolute power by 1927. I will not accuse Koestler
+himself of assuming that Lenin was violent but just whereas Stalin was unjust and violent,
+but Rubashov appears to believe that, and Ponty certainly does. This perspective was
+widespread among both Soviets and global leftists, but must be questioned. As I'm sure
+you've by now guessed, the counterpoint I find most striking to Ponty and Rubashov's ideal
+of Lenin is Lenin's 1918 telegraph ordering the hanging of 100 landed farmers. Lenin uses
+the word \enquote{kulak} derogatorily to describe the landed farmers who opposed
+collectivization, and I'll freely admit that based on context clues I first assumed that
+the \textit{kulaki} were an ethnic group. In fact, the \textit{kulaki} were a sort of
+middle class in tsarist Russia, though not the city-dwelling petty bourgeoisie described
+by Marx. Lenin's descriptions of the \textit{kulaki} are absurdly violent; he called them,
+\textquote[{\cite[65]{rubinstein}}]{bloodsuckers, vampires, plunderers of people and
+profiteers, who fatten on famine}. Though the \textit{kulaki} were better off than
+\textit{bednjaki}, they were far from rich; they might have owned a couple head of cattle
+or a few acres of land \parencite[94]{conquest}. Still, their refusal to give up crops to
+the new government was enough for Lenin to order,
 
-At this point in the paper, I had intended to use textal evidence from \lear{}
-to convince you that Lear's mental deterioration caused the war between the
-English and the French, but honestly, it seems unnecessary. That reading is so
-obviously on the surface of this play that I won't waste either of our time
-beating a dead horse.
+\blockquote[{\cite{lenin}}]{The revolt by the five kulak volost's must be suppressed
+without mercy. \textelp{} We need to set an example.
 
-% \subsection*{Blindness}
+\begin{enumerate}
+\item You need to hang (hang without fail, so that the public sees) at least 100 notorious
+kulaks, the rich, and the bloodsuckers.
 
-% \section*{Tell 'em what you told 'em}
+\item Publish their names.
 
-Shakespeare's \lear{} and Sophocles' \oedipus{} both present the dangers
-inherint in absolute rule: the ruler is placed under incredible stress, and a
-ruler who succumbs to stress will cause great harm not just to themself but to
-the state and all its citizens. There are, in my mind, two potential responses
-to this problem, though neither author explicitly suggests either. In a
-monarchist reading, the consequence is that kings (and tyrants) must be
-exceptionally steady, level-headed people, and must be surrounded by strong,
-trusted advisors who are empowered to stop the ruler from making poor
-choices. Both plays show a hurdles in this solution: Lear and Oedipus are both
-professed to be the steady, level-headed type, and both are surrounded by
-trusted advisors; neither of these safeguards prevents the tragedy of the plays
-from unfolding. The antimonarchist reading is, of course, that divesting
-absolute rule in a single person is always a losing proposition, no matter how
-well-suited that ruler appears, and so political systems which distribute power
-among multiple people must be preferred.
+\item Take away all of their grain.
+
+\item Execute the hostages -- in accordance with yesterday's telegram.
+\end{enumerate}
+
+This needs to be accomplished in such a way, that people for hundreds of miles around will
+see, tremble, know and scream out: let's choke and strangle those blood-sucking kulaks.}
+
+I do not live in Ponty's \textit{Stimmung} of violence, and so by his standards I am
+unequipped to criticize Lenin's decision to publicly, violently execute dissenters with
+the explicit goal of inspiring his followers to further violence. More importantly, I am
+disinclined to engage in the sort of historical speculation that follows from asking,
+\enquote{what might have happened if Lenin had not responded so violently to the
+\textit{kulaki}? Might some of their number been swayed by promises of a better life,
+rather than fear?} No use will come of asking these questions, because we can never know
+the answers. We do know for certain, though, that in this historical moment, Lenin was
+encouraging violence not against imperialists and capitalists living idly in palaces
+supported by throngs of malnourished laborers, but against hardworking farmers who were a
+bit further from starvation than their neighbors, objecting to a state seizure of land and
+food which would have -- and did -- leave them unable to feed themselves or their
+families. And we also know, looking back, that Lenin's choices were a step down the path
+of a state whose violence did not \textquote[{\cite[xviii]{ponty}}]{recede\textdel{s} with
+the approach of man's future}, as Ponty said it should have. This hardly seems to have the
+sense Ponty attributes to his idealized Lenin.
+
+%%% stimmung of violence; this is just a cop-out which prevents criticizing violence
+
+%% resolve section
+Ponty's insistence that we critique Soviet violence as a divergence between Soviet dogma
+and Marxist ideology is flawed for two reasons: firstly, because Koestler's book is a
+critique of the violence in Soviet dogma, not the violence in Marxist ideology; and
+secondly, because Ponty refuses to criticize the choices which led Soviet revolutionaries
+to construct their dogma.

### 2022-04-25 paper.tex
initial draft of my final project for "reimagining greek tragedy"

--- 2022-03-18 paper.tex
@@
-\newcommand{\lear}{\textit{King Lear}}
-\newcommand{\oedipus}{\textit{Oedipus Tyrannus}}
-% \section*{Tell 'em what you're gonna tell 'em}
-Shakespeare's \lear{} and Sophocles' \oedipus{}, despite being separated by two
-millenia and situated in vastly different political systems, show similar
-outlooks on the health and fitness of the monarch, of the ruling family, and of
-the state. Both plays are concerned with the feedback loop between the stress of
-rulership, the monarch's mental health, and the disastrous ripples caused by the
-monarch's short-sighted decision-making. Sophocles' more reserved, Aristotelian
-dramatic form embodies these concerns almost exclusively in Oedipus himself,
-while Shakespeare's greater complexity allows him to spread the embodiments of
-illness between Lear and Gloucester. Sophocles' style is more ambiguous,
-inviting the audience to empathize with Oedipus even as we watch him descend
-into madness and take Thebes with him, whereas Shakespeare is content to have us
-dislike Lear, with the tragedy of the play coming not from his personal
-suffering but the suffering he inflicts on those around him. Still, the two
-plays show remarkable consistency in their depictions of impulsivity,
-self-assuredness and rage and their linking of those traits to the suffering and
-downfall of the state.
+\newcommand{\theplay}{\textit{Oedipus Tyrannus}}%
+\newcommand{\thegame}{\textit{Oedipus Text} (working title)}%
+% \section*{Background}
 
-% \section*{Tell 'em}
+% \subsection*{The source text}
 
-% \subsection*{Impulsivity, self-assurance, rage}
+\theplay{} is the most widely-read and widely-known of the Greek tragedies. In
+\textit{Poetics}, Aristotle lifts it up as the epitome of the form. \footnote{Why
+  Aristotle chooses \theplay{} as the greatest tragedy, why he chooses tragedy as the
+  greatest medium, and how \theplay{} compares in form and content to other Greek tragedies
+  are all questions for another day; In this project I am concerned only with the fact
+  that \theplay{} is, for all intents and purposes, the canonical Greek tragedy.} Because it
+is widely taught and widely referenced in later culture and media, \theplay{} presents a
+relatively accessible first source text for a new mode of presenting Greek tragedy.
 
-At the beginning of \oedipus{}, the titular tyrant is quite lucid; it is only as
-the play progresses and his fate becomes clear that his mental state is called
-into question. He first suffers a burst of rage when Tireseas refuses to reveal
-the identity of Laius' killer; Oedipus' inflamed response is,
-\textquote[{\cite[ll. 334--335, 339--340]{oedipus}}]{Nothing? Damn you! You
-  could make the coldest stone // Burn with rage. \textelp{} Could anyone not be
-  angry after hearing // How you hold our city in such contempt?} Though
-Tireseas is confoundingly unhelpful, he is not wrong to refuse Oedipus' demands,
-nor is Oedipus wrong to be enraged by them. Sophocles shows us that Oedipus, in
-the course of his duties as tyrant, is unavoidably exposed to circumstances
-which are stressful beyond human tolerance.
+Sophocles' text is quite deep, and adaptors must select from among its themes and facets
+what to focus on. In descending order of my interest, I will engage with:
 
-Lear, on the other hand, has already succumbed to that stress by the time his
-play begins. Instead of watching his descent from lucidity to madness, we are
-forced to reconstruct an image of the king he once was by contrasting his older
-advisors' loyalty to him against the younger generation's eagerness to oust
-him. Kent, after being banished and threatened with death for opposing Lear
-\parencite[1.1]{lear}, returns in disguise to continue his service to the
-king. Lear is justifiably confused that an apparent stranger would offer himself
-into the service of a destitute monarch, and the audience is similarly confused
-that Kent would return to his thankless and dangerous work. In an aside, Kent
-explains to us, \textquote[{\cite[1.4]{lear}}]{thy master, whom thou lovest, //
-  Shall find thee full of labours}. To Lear, Kent gives us a glimpse of the
-traits which inspired that love, swearing \textquote[{\cite[1.4]{lear}}]{to
-  serve him truly that will put me in trust: to love him that is honest; to
-  converse with him that is wise \textelp{}}. A somewhat contrived turn of
-dialogue has Lear ask Kent his age, and we are told that Kent is 48. This scene
-shows us Lear's former glory, the years Kent spent in service of an honest,
-wise, just king. We cannot help but be struck by the contrast between Kent's
-description and the Lear who stands before us onstage.
+\paragraph*{The tension between destiny and self-determination.} Is the oracle's prophecy
+inevitable, or could Oedipus have taken a different path? This is complicated by the
+dramatic form; the script written by Sophocles acts as its own prophecy, and no matter how
+many times \theplay{} is re-enacted onstage, the outcome will be the same.
 
-Goneril draws our attention to the transformation again later in the same scene,
-entreating her father, \textquote[{\cite[1.4]{lear}}]{I would you would make use
-  of that good wisdom \textelp{} and put away // These dispositions, that of
-  late transform you // From what you rightly are}. Lear, of course, objects,
-insisting that he is now as he's always been, but we know that is not the
-case. The stress of rule has taken its toll on Lear, and he is no longer the man
-he once was.
+\paragraph*{Rage inhibiting rational decision-making,} or the tension between Freud's
+\textit{it} and \textit{I}. The \textit{hamartia} which sets the tragedy in motion and
+ensures the prophecy is Oedipus' loss of self-control when provoked at the
+crossroads. Could another man have kept their cool in that moment? This is again
+complicated by the nature of the play; Oedipus has to lose control in that moment and kill
+Laius, or else there is no story for Sophocles to tell.
 
-The symptoms displayed by Lear and Oedipus bear highlighting: both are prone to
-paranoia and fits of rage. Oedipus sees schemes against himself, Thebes and
-Laius around every corner; he accuses Tireseas of having part in Laius' death
-\parencite[ll. 345--349]{oedipus}, Creon of conspiring to falsely accuse Oedipus
-and claim power for himself \parencite[ll. 385--389, 399--402,
-572--573]{oedipus}, and espouses a philosophy of rule based in mutual
-conspiracy, \textquote[{\cite[ll. 618--621]{oedipus}}]{When a conspirator moves
-  quickly against me, // Then I must be quick to conspire back at him. // If I
-  hesitate I will lose the initiative, // While he will seize the moment and
-  strike}. Sophocles, relentlessly didactic, does not place the onus on us to
-interpret and reject Oedipus' paranoia, but has the chorus tell us,
-\textquote[{\cite[ll. 616--617]{oedipus}}]{Anyone who treads carefully will
-  agree: // Decisions made too quickly are dangerous}. Oedipus then demands that
-Creon (who is portrayed in \textit{oedipus} as noble, just and content, with no
-lust for power) be put to death, leading to a fast-paced dialogue revelatory of
-Oedipus' state:
+\paragraph*{Fitness to rule and a monarch's responsibility to their title.} I investigated
+this theme in detail in my paper \enquote{Body and the state: Illness in \textit{Oedipus
+    Tyrannus} and \textit{King Lear}}, but in short, Oedipus is a good guy whose only flaw
+is his tendency to lose control of his rage when provoked. His temper is certainly not a
+good thing, but it's fundamentally understandable, and well within the bounds of normal
+behavior for a classical Greek nobleman. But still, as \theplay{} shows us, it makes him
+unfit to rule Thebes. If a wise, caring, righteous man whose reactions to stressful
+circumstances are within social norms is not a fit monarch, is anyone?  \footnote{Spoiler
+  for my paper: no, no individual is fit to rule a society.}
 
+% \subsection*{Interactive Fiction}
+
+\paragraph*{Choose-your-own-adventure games} are a genre of Interactive Fiction (IF), or
+text-based computer games. Classic IF games like \textit{Zork} used a user-interaction
+model where the player typed commands, like \texttt{open door}, \texttt{walk north} or
+\texttt{take key}. I feel a certain nostalgia for these games \footnote{Despite having
+  been born long after their time\ldots{}}, but they're often frustrating, as the player
+has no way of knowing what commands the game will accept. The genre is also known for
+oblique puzzles, which I believe tend to distract from the narrative.
+
+More modern narrative text-based games tend instead to use the Choose-Your-Own-Adventure
+(CHYOA) model, where relatively long blocks of text are separated by points of
+interaction. I will call these points of interaction \enquote{dilemmas,} though that term
+is not in the common parlance. At each dilemma, the player chooses between a small number
+of options offered by the writer. Subsequent blocks of text then reflect the player's
+choice.
+
+When developing CHYOA games, writers are faced with a trade-off between player agency and
+implementation effort. On the one hand, the whole point of interactive fiction is to allow
+the reader to influence the outcome of the story. Players tend to desire a relative
+frequency of dilemmas and a relative abundance of options, and for choices which lead to
+meaningfully different outcomes. If a writer's goal was just to tell a singular story that
+followed a particular path and ended in a predetermined way, they would eschew the
+interactive element and write a traditional narrative in one of the many modes available.
+
+On the other hand, currently extant tools for writing CHYOA games require that the writer
+enumerate all possible choices and write the text for all possible outcomes. This is, to
+put it lightly, a lot of work. And writers who invest too much effort in offering many
+options which lead to diverse outcomes will find that each of those outcomes is less
+polished than if they had focused on a smaller number of better-developed paths. Also,
+players will often explore only a single path; the prospect of writing one hundred words
+of which each reader will experience only ten is creatively unsatisfying.
+
+All narrative computer games grapple with this tension to some extent. \textit{Mass Effect
+  3}, the final entry in a high-budget trilogy of role-playing shooter games which
+advertised the player's influence over the story, was famously criticized because all
+paths converged on similar endings. \textit{Katana Zero}, an independent hack-and-slash
+Vietnam War allegory, explored the tension by forcing players to choose between a
+\enquote{good} ending which ended the game immediately and a \enquote{bad} ending which
+featured significantly more content, at the cost of making the player feel guilty for
+actively continuing the game's violence. But the problem is most pronounced in CHYOAs due
+to their lack of any gameplay beyond the story. \textit{Black Mirror: Bandersnatch},
+Netflix's thoroughly metatextual CHYOA film about the developer of a CHYOA game, is
+explicit in its attempt to grapple with the problem.
+
+% \section*{Motivation}
+
+An adaptation of \theplay{} as a CHYOA game provides a unique opportunity to wed
+Sophocles' tension between self-determination and fate with the CHYOA writer's tension
+between branching paths and finite scope. To that end, I propose \thegame{}. As my first
+adaptation of a Greek tragedy into this mode, and, to the best of my knowledge, the first
+explicit adaptation of a classical text as a computer game, \thegame{} will necessarily
+also serve as a design manifesto.
+
+% \section*{Design}
+
+\newcommand{\selfcontrol}{\texttt{\textsc{Self Control}}}
+
+\newcommand{\knowledge}{\texttt{\textsc{Knowledge}}}
+
+\paragraph*{\thegame{} will be a mystery, where the player attempts to assemble enough
+  \knowledge{} to discover the cause of the plague.} As the player traverses the story,
+they will uncover facts, which will appear in their inventory. For example, during their
+interview with the shepherd, the player will learn \texttt{Laius and Jocasta birthed a
+  son, but ordered it put to death by leaving it on a hillside}. Further investigating the
+shepherd will add, \texttt{Laius and Jocasta's baby was not put to death, but handed off
+  to a Corinthian}. Not all facts will be reliable, and incorrect facts may at times be
+replaced with corrected versions; \texttt{I am Oedipus, son of Polybus and Merope, prince
+  of Corinth} will eventually be replaced with \texttt{Polybus and Merope adopted me and
+  raised me as their son}, and \texttt{Laius was killed by a band of thieves. Plural} with
+\texttt{Laius was killed by a lone traveler}.
+
+Completing the game will require the player to accrue enough \knowledge{} to
+incontrovertibly prove that Oedipus is the son of Laius and Jocasta and the killer of
+Laius.
+
+\paragraph*{The core mechanic of \thegame{} will be \selfcontrol{}, } a finite resource
+which can be spent to make better choices. In the first draft of the game, it is
+described,
+
+\newcommand{\gamequote}[1]{
   \begin{minipage}[t]{0.8\linewidth}
-  \begin{dialogue}
-    \speak{Creon} And you could be wrong.
-    \speak{Oedipus} Nevertheless, I must rule.
-    \speak{Creon} Not if it means ruling badly.
-    \speak{Oedipus} Oh, my city, my city\ldots{}
-    \speak{Creon} I belong in this city, it's not only yours!
+    \texttt{#1}
+  \end{minipage}
+  \vspace{\baselineskip}
+}
 
-    \attrib{\cite[ll. 128--130]{oedipus}}
-  \end{dialogue}
-\end{minipage}
+\gamequote{\textsc{Self Control} quantifies your ability to suppress your impulses and
+  make good choices. Spend it wisely; fighting your nature and controlling your urges is
+  stressful. If you lose your \textsc{Self Control}, you may find yourself making
+  impulsive choices you'll end up regretting.}
 
-\vspace{0.5\baselineskip}
+At each dilemma, some number of options will be available which do not cost any
+\selfcontrol{}. These will tend to be more impulsive actions. Some number of options will
+also be available which cost various amounts of \selfcontrol{}. These will tend to be more
+rational, thought-out actions.
 
-In this moment, we see that the stresses of ruling (and espeically, of trying to
-fix the plague caused by his rule) have driven Oedipus from a just tyranny to
-the violence and oppression described by Plato. Oedipus feels pressured to make
-quick, firm decisions even when they are wrong, and sees the city as his
-possession rather than the object of his service. What is worse, we do not hate
-Oedipus for this denigration any more than Kent hates Lear; it is the natural
-consequence of circumstances outside Oedipus' control.
+Consider as an example the scene where Creon warns Oedipus to believe Tiresias and not to
+act impulsively. \footnote{I have not written this scene yet, so what follows is far from
+  final.} The player would then face a dilemma with two available options:
 
-% \subsection*{Suffering and downfall of the state}
+\begin{enumerate}
+\item At no cost in \selfcontrol{}, the player could re-enact Sophocles' text by choosing
+  to yell at Creon and threaten him with exile.
+\item In exchange for one \selfcontrol{}, the player could accept Creon's critique
+  graciously and apologize to Tiresias. This would have some yet-undetermined benefit to
+  the player, possibly granting them a piece of \knowledge{} which would aid them in
+  decoding the mystery.
+\end{enumerate}
 
-Both \lear{} and \oedipus{} are keen to show us that, because of their
-responsibilities to their respective states, the tragic suffering of the titular
-rulers extends to their subjects. For Oedipus, this manifests as a plague, while
-for Lear it is a war. (I am tempted to investigate these different portrayals of
-danger to the state in search ofsome statement about the political circumstances
-of Athens in 429 BCE and England in 1606, but that seems at best tangentially
-related to the argument I'm making here.) Again, the structures of these plays
-lead us to experience the suffering differently; Sophocles' reliance on a
-well-known myth, his limited scope and his character-driven focus places us
-already in the plague at the start of the play, without depicting its
-beginnings. We are shown, however, the gravity and horror of the plague
-\parencite[e.g. ll. 22--30]{oedipus}, and we are reminded that the plague is a
-divine punishment for Oedipus' hamartia, his killing of Laius
-\parencite[ll. 101--110]{oedipus}. Much later, Sophocles eplicicitly links that
-killing to Oedipus' temper \parencite[ll. 807--813]{oedipus}. After, quite
-frankly, a very long journey, our understanding of events is: Oedipus, in a
-(provoked and arguably justified) fit of rage, killed Laius, then, without
-realizing, claimed his throne and house. The gods, outraged that Thebes would,
-also without realizing, accept the killer of their rightful king and crown him
-tyrant, strike the city with a plague. All of which is to say that the Theban
-plague was caused by Oedipus' loss of control over his temper.
+Crucially, the player's supply of \selfcontrol{} will be constrained such that it is
+impossible to reach any ending except the one we know and love from the myth, the poetry
+and \theplay{}. Key pieces of \knowledge{}, which, when combined, could lead Oedipus down
+a better path, will be strategically locked behind options which, in total, consume more
+\selfcontrol{} than the player has access to. This way, players who revisit to \thegame{}
+and make different choices in search of a better ending will find that Oedipus' rage
+prevents them from doing so.
 
-% lear causes a war, lmao
+At key story-defining moments, most notably killing Laius, the player will be presented
+with options which cost more \selfcontrol{} than the player's entire reserve. These
+moments will always be associated with an injury to Oedipus' feet. The player will
+understand that Oedipus' (shame about) his disability
 
-At this point in the paper, I had intended to use textal evidence from \lear{}
-to convince you that Lear's mental deterioration caused the war between the
-English and the French, but honestly, it seems unnecessary. That reading is so
-obviously on the surface of this play that I won't waste either of our time
-beating a dead horse.
+\paragraph*{In order to serve its goals as an exemplar of its new mode,} \thegame{} will
+be fiercely and unapologetically metatextual. In addition to narration which will not be
+attributed to any speaker, the text will feature design notes in the form of dialogue
+attributed to myself, \texttt{phoebe}. In these moments, I will describe to the player my
+design choices and intentions, instruct the player as to the game's mechanics and how to
+interact with them, and explain background information about the setting. When Oedipus is
+crowned Tyrant of Thebes, for example, the player is offered an opportunity to object,
+\texttt{Wait, tyrants are bad. I don't want to be a tyrant. Can't I be the King of
+  Thebes?} If they do, I respond,
 
-% \subsection*{Blindness}
+\gamequote{\textsc{phoebe}:
+  \begin{minipage}[t]{0.8\linewidth}
+    Great question! Tough to answer without a lesson in classical Greek civics, and I
+    don't want to get to bogged down in that. But short answer: no, tyrants aren't
+    bad. Well, maybe they are, but the Greeks haven't caught on to that yet. In this
+    world, a tyrant is someone who takes rulership of a city through might, or through
+    cunning, or popular support, or who steps up to lead the people through a crisis. Like
+    you did with the Sphinx. A king, on the other hand, is someone who inherits rule from
+    their father, or who is appointed by the gods. If you had been the son of Laius, you
+    would've inherited kingship of Thebes; but you got the throne in a crisis, so you're a
+    tyrant.
+\end{minipage}}
 
-% \section*{Tell 'em what you told 'em}
+From there, if they player has not yet uncovered the truth of their parentage, they may
+say, \texttt{I wish I was Laius' son. Then I'd be a real king,} to which I reply,
+\texttt{Yeah, too bad, huh? If only.} If the player does know that Oedipus is the son of
+Laius, then they may instead say, \texttt{But wait, I am the son of Laius!} to which I
+reply, \texttt{Isn't that ironic?}
 
-Shakespeare's \lear{} and Sophocles' \oedipus{} both present the dangers
-inherint in absolute rule: the ruler is placed under incredible stress, and a
-ruler who succumbs to stress will cause great harm not just to themself but to
-the state and all its citizens. There are, in my mind, two potential responses
-to this problem, though neither author explicitly suggests either. In a
-monarchist reading, the consequence is that kings (and tyrants) must be
-exceptionally steady, level-headed people, and must be surrounded by strong,
-trusted advisors who are empowered to stop the ruler from making poor
-choices. Both plays show a hurdles in this solution: Lear and Oedipus are both
-professed to be the steady, level-headed type, and both are surrounded by
-trusted advisors; neither of these safeguards prevents the tragedy of the plays
-from unfolding. The antimonarchist reading is, of course, that divesting
-absolute rule in a single person is always a losing proposition, no matter how
-well-suited that ruler appears, and so political systems which distribute power
-among multiple people must be preferred.
+\paragraph*{The plot of \thegame{}} will be expanded in scope relative to \theplay{}, in
+an effort to expose more of Oedipus' backstory and character. This feels like an obvious
+choice, as narrative prose is not subject to the same forces which incentivize drama
+towards smaller scopes and shorter plots. \footnote{I generally think Aristotle is right
+  that tragedies (and all plays) should be small in scope, but that's neither here nor
+  there.} In service of allowing the player to understand and inhabit Oedipus, \thegame{}
+will follow him from his childhood in Corinth. The rough outline of the scenes I have
+planned are:
+
+\begin{enumerate}
+\item In Corinth, a short scene of Oedipus' childhood will introduce his disability, his
+  related insecurity, and his temper via a schoolyard scuffle.
+\item As a young adult, a tutorial set in Polybus' court will introduce the game's
+  mechanics; namely facing dilemmas, evaluating options, making choices, and balancing
+  spending \selfcontrol{}.
+\item At a banquet at Corinth, Oedipus will overhear a rumor about his prophecy from a
+  drunk guest. The player will then choose between spending \selfcontrol{} to ignore the
+  rumor and stay in Corinth, or running away to Delphi to confirm the rumor. If the player
+  opts to stay, the game will end with some lines about how, on the one hand, Oedipus has
+  escaped his fate, but on the other hand, there isn't any story to tell. The player will
+  be encouraged to go back and replay, choosing to leave for Delphi.
+\item At Delphi, Oedipus will here the prophecy. Unlike the rest of the game's text, the
+  oracle will speak in verse, possibly using lines taken from Gilbert Murray's 1911
+  translation in rhyming verse. \footnote{The oracle does not deliver the prophecy in
+    \theplay{}, but it contains enough attestations of the prophecy that I think I could
+    piece something together.} No opportunity for player interaction will be provided
+  during this scene.
+\item At the foot of Mount Parnassus, Oedipus will be insulted by a carriage-driver, who
+  will whip at Oedipus' feet to get him to move out of the way. An option will be shown to
+  have Oedipus keep his cool, with a description that calls out how unwise it is to get
+  into fights when one is trying to avoid a prophecy about committing murder, but it will
+  be programmed to require one more \selfcontrol{} than the player has at the moment,
+  regardless of how much they have spent in preceding scenes. The only option actually
+  available to the player will be described simply as \texttt{Lose your cool}, and will
+  describe Oedipus' slaughtering the entire carriage procession.
+\item After Oedipus comes to his senses, the player will be offered a chance, at the cost
+  of \selfcontrol{}, to return to Corinth in spite of the prophecy. This will result in an
+  ending similar to the one if the player stays in Corinth after hearing the
+  rumor. Otherwise, the player will continue on to Thebes.
+\item The player will confront the Sphinx at Thebes. Dialogue attributed to me will call
+  out the Sphinx's tie to Egypt, and invite the player to reflect on it. The player will
+  be shown a text box and asked to type the answer to the Sphix's riddle,
+  \texttt{man}. After each incorrect guess, dialogue attributed to me will offer
+  increasingly obvious hints, including \texttt{Just go Google \enquote{riddle of the
+      sphinx,} okay?}
+\item Joyful citizens of Thebes will crown Oedipus Tyrant of Thebes.
+\item A series of vignettes will show a few years of Oedipus' reign as tyrant, the
+  citizens' satisfaction with his rule, his marriage to Jocasta and the birth of his
+  children.
+\item A final vignette will introduce the plague, and Creon will arrive to inform the
+  player of the curse and the requirements for dispelling it.
+\item Oedipus' investigation will play out non-linearly, with the player choosing the
+  order to investigate various leads. After pursuing a lead, they will be returned to a
+  hub scene set at the palace at Thebes, where they can choose another.
+\item Once the player has accumulated enough \knowledge{} to solve the mystery
+  \footnote{Or, once they have accumulated enough \knowledge{} that they can no longer
+    deny the obvious.}, they will choose between killing themselves, or, at the cost of
+  \selfcontrol{}, putting out their eyes and leaving the city. Players who spend too much
+  \selfcontrol{} during the investigation will have no option but suicide. If the player
+  chooses suicide, the game will end immediately, with no resolution. Otherwise,
+  \thegame{} will end the same way as \theplay{}.
+\end{enumerate}
